---
title: "Getting Started with Rwapor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Rwapor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The **Rwapor** package provides tools for downloading and processing WaPOR
(Water Productivity Open-access portal) and AgERA5 climate data from the
FAO GISMGR API. This vignette covers the basic workflow for:

1. Downloading raster maps for a region
2. Extracting time series with zonal statistics
3. Converting units between temporal resolutions

## Installation

```{r installation}
# Install from GitHub
# install.packages("pak")
pak::pak("almutaz9000/Rwapor")

library(Rwapor)
```

## Performance: GDAL & Batching

Rwapor automatically optimises GDAL's HTTP streaming settings when the
package loads. This dramatically reduces download times for
Cloud-Optimized GeoTIFFs (COGs) by using 10 MB read chunks instead of
GDAL's 16 KB default, cutting the number of HTTP round-trips by 100–1000×.

### Automatic Spatial Fixes (PROJ Redirection)

On Windows systems with multiple GIS installations (e.g., PostGIS via PostgreSQL),
the `PROJ_LIB` environment variable can point to incompatible databases, causing
coordinate system errors. Rwapor automatically detects and redirects these to its
internal, compatible database when the package loads:

```r
library(Rwapor)
# This calls wapor_fix_proj() automatically
```

### Optional Batching & Parallel Processing

For long time series (e.g., 5+ years of dekadal data), loading all rasters at 
once can exceed available RAM. Both `wapor_map()` and `wapor_ts()` support 
**optional batching** and **parallel processing** to manage resources:

```r
library(Rwapor)
library(future)

# 1. Setup parallel backend (optional)
plan(multisession, workers = 4)

# 2. Extract with parallel batching
# This will split 36 layers into 12 batches of 3, processed in parallel.
df_batched <- wapor_ts(
  region     = c(35.0, 33.0, 36.0, 34.0),
  variable   = "L1-AETI-D",
  period     = c("2023-01-01", "2023-12-31"),
  batching   = TRUE,
  batch_size = 3,
  parallel   = TRUE
)
```

- **`batching`**: Defaults to `TRUE`. Processes data in manageable chunks.
- **`batch_size`**: Number of layers per chunk (default `12L`).
- **`parallel`**: If `TRUE`, batches are processed in parallel via background workers.

## Available Variables

### WaPOR Variables

WaPOR provides water productivity and related variables at different spatial
and temporal resolutions:

```{r wapor-vars}
# List all WaPOR v3 variables
names(WAPOR3_VARS)

# Get metadata for a specific variable
get_variable_metadata("L1-AETI-D")
# $long_name: "Actual EvapoTranspiration and Interception"
# $units: "mm/day"
# $scale: 0.1
```

**Variable naming convention:** `{Level}-{Variable}-{TemporalResolution}`

- **Levels:** L1 (continental ~250m), L2 (regional ~100m), L3 (local)
- **Variables:** AETI, E, I, NPP, PCP, T, GBWP, NBWP
- **Temporal:** D (dekadal), M (monthly), A (annual), E (daily)

### AgERA5 Climate Variables

```{r agera5-vars}
# List AgERA5 variables
names(AGERA5_VARS)

# Example: Reference evapotranspiration
get_variable_metadata("AGERA5-ET0-E")
```

## Downloading Raster Maps

Use `wapor_map()` to download and save raster data for a region:

### Using a Bounding Box

```{r download-bbox}
# Define region as bounding box: c(xmin, ymin, xmax, ymax)
region <- c(35.0, 33.0, 36.0, 34.0)

# Download dekadal ET for January 2023 (defaults to mm/dekad)
output_file <- wapor_map(
  region   = region,
  variable = "L1-AETI-D",
  period   = c("2023-01-01", "2023-01-31"),
  folder   = "output"
)

print(output_file)
# "output/L1-AETI-D/bb_WAPOR-3.L1-AETI-D.2023-01-01_2023-01-21.tif"
```

### Using a Shapefile

```{r download-shapefile}
# Download for a shapefile region, cropped and masked to the polygon
output_file <- wapor_map(
  region   = "path/to/my_region.shp",
  variable = "L2-NPP-M",
  period   = c("2023-01-01", "2023-12-31"),
  folder   = "output",
  filename = "npp_monthly_2023.tif"
)
```

### Using L3 Region Codes

For WaPOR Level 3 data, use the 3-letter region code:

```{r download-l3}
# Download for Awash Basin (Ethiopia)
output_file <- wapor_map(
  region   = "AWA",
  variable = "L3-AETI-D",
  period   = c("2023-06-01", "2023-06-30"),
  folder   = "output"
)
```

### Saving Separate Files Per Time Step

```{r download-separate}
# One GeoTIFF per dekad instead of a single multi-band stack
paths <- wapor_map(
  region         = c(35.0, 33.0, 36.0, 34.0),
  variable       = "L1-AETI-D",
  period         = c("2023-01-01", "2023-03-31"),
  folder         = "output",
  separate_files = TRUE
)

length(paths)  # 9 files (3 dekads × 3 months)
```

### Multiple Variables

```{r download-multi}
# Download ET and NPP in one call — returns a named list
results <- wapor_map(
  region   = c(35.0, 33.0, 36.0, 34.0),
  variable = c("L1-AETI-D", "L1-NPP-D"),
  period   = c("2023-01-01", "2023-03-31"),
  folder   = "output"
)

results[["L1-AETI-D"]]
results[["L1-NPP-D"]]
```

## Extracting Time Series

Use `wapor_ts()` to extract zonal statistics (mean, min, max) for polygons:

### Basic Time Series Extraction

```{r ts-basic}
# Extract time series for a bounding box (defaults to mm/dekad for dekadal vars)
df <- wapor_ts(
  region   = c(35.0, 33.0, 36.0, 34.0),
  variable = "L1-AETI-D",
  period   = c("2023-01-01", "2023-03-31")
)

head(df)
#   mean   min   max start_date   end_date  number_of_days layer_index
# 1  2.5   1.2   4.1 2023-01-01 2023-01-10             10           1
# 2  2.8   1.5   4.3 2023-01-11 2023-01-20             10           2
```

### Time Series for Multiple Polygons

```{r ts-polygons}
# Extract for multiple fields — identifier column names each polygon
df <- wapor_ts(
  region     = "fields.geojson",
  variable   = "L1-AETI-D",
  period     = c("2023-01-01", "2023-12-31"),
  identifier = "field_name"
)

head(df)
```

### Accessing Metadata

The returned data frame carries units and variable description as attributes:

```{r ts-metadata}
attr(df, "units")
# "mm/dekad"

attr(df, "long_name")
# "Actual EvapoTranspiration and Interception"
```

## Automatic Data Transformations

To streamline agricultural water productivity and crop monitoring, `Rwapor` automatically performs key unit and variable transformations during the download and extraction processes.

### 1. WaPOR Dekadal Defaults (Daily Rates to Dekadal Totals)

WaPOR dekadal (10-day) variables are published on the FAO portal as average daily rates (`mm/day` or `gC/m²/day`). However, crop and water productivity analyses typically require total water depth or biomass produced over each 10-day period.

To eliminate manual date-length calculations:
* By default, `wapor_ts()` and `wapor_map()` **automatically scale and convert daily rates to dekadal totals** (`mm/dekad` or `gC/m²/dekad`) based on the exact number of days in each specific dekad (either 10, or 11/8/9 depending on the month and leap year).
* You can customize or disable this conversion using the `unit_conversion` parameter (e.g., set `unit_conversion = "day"` to retrieve daily rates, or `"month"` to aggregate to monthly totals).

```{r unit-conversion}
# Default: mm/dekad (dekadal total)
df_dekad <- wapor_ts(
  region   = c(35.0, 33.0, 36.0, 34.0),
  variable = "L1-AETI-D",
  period   = c("2023-01-01", "2023-03-31")
)
attr(df_dekad, "units")  # "mm/dekad"

# Retrieve daily rate (mm/day)
df_day <- wapor_ts(
  region          = c(35.0, 33.0, 36.0, 34.0),
  variable        = "L1-AETI-D",
  period          = c("2023-01-01", "2023-03-31"),
  unit_conversion = "day"
)
attr(df_day, "units")          # "mm/day"
attr(df_day, "original_units") # "mm/day"

# Monthly total
df_monthly <- wapor_ts(
  region          = c(35.0, 33.0, 36.0, 34.0),
  variable        = "L1-AETI-D",
  period          = c("2023-01-01", "2023-03-31"),
  unit_conversion = "month"
)
attr(df_monthly, "units")  # "mm/month"
```

Available conversions: `"none"`, `"day"`, `"dekad"`, `"month"`, `"year"`

### 2. Automatic Temperature Conversion (Kelvin to Celsius)

AgERA5 temperature datasets (`AGERA5-TMIN-E`, `AGERA5-TMAX-E`) are stored natively on the FAO Copernicus mirrors in Kelvin (`K`). To make the data immediately usable for agricultural and phenological models:
* Both `wapor_map()` and `wapor_ts()` **automatically convert Kelvin values to degrees Celsius (°C)** by subtracting `273.15` during raster download and extraction.
* The output SpatRaster layers and extracted statistical data frames are in Celsius, and their units metadata attribute is dynamically updated from `K` to `degC` or `°C` automatically.

## Working with URLs Directly

For advanced use cases, generate download URLs without saving to disk:

```{r generate-urls}
# Get download URLs for a variable and period
urls <- wapor_generate_urls(
  variable = "L1-AETI-D",
  period   = c("2023-01-01", "2023-01-31")
)

length(urls)  # 3 (one per dekad)

# Stream directly with terra — GDAL settings applied by Rwapor ensure
# these remote reads are fast
library(terra)
r <- terra::rast(paste0("/vsicurl/", urls))
```

## Parsing Regions

The `parse_region()` function normalises different region formats:

```{r parse-region}
# Bounding box
parse_region(c(35.0, 33.0, 36.0, 34.0))
# $type: "bbox"

# L3 region code
parse_region("AWA")
# $type: "l3_code"

# Vector file
parse_region("path/to/region.shp")
# $type: "vector"
```

## Error Handling

All functions provide informative error messages:

```{r error-handling, error = TRUE}
# Invalid variable format
wapor_generate_urls("INVALID")
# Error: Invalid variable format 'INVALID'. Expected format: 'LEVEL-VAR-TRES'

# Invalid date range
wapor_generate_urls("L1-AETI-D", period = c("2023-12-31", "2023-01-01"))
# Error: Start date must be before or equal to end date

# Invalid bounding box
parse_region(c(36, 33, 35, 34))
# Error: Invalid bounding box: xmin must be less than xmax
```

## Best Practices

1. **GDAL is tuned automatically** — `library(Rwapor)` applies optimised HTTP
   settings immediately. No extra setup required. Call `wapor_gdal_settings()`
   to confirm or `wapor_configure_gdal()` to customise.

2. **Check variable metadata** before downloading to understand units:
   ```r
   get_variable_metadata("L1-AETI-D")
   ```

3. **Use appropriate temporal resolution** for your analysis:
   - Daily (E): Weather / irrigation scheduling
   - Dekadal (D): Crop monitoring
   - Monthly (M): Seasonal analysis
   - Annual (A): Long-term trends

4. **Use `unit_conversion`** to get the right aggregation level:
   - `"dekad"` — 10-day total (default for dekadal variables)
   - `"month"` — monthly total
   - `"day"`  — daily rate

5. **Use `separate_files = TRUE`** in `wapor_map()` when you need one file
   per time step, for example to feed into external processing pipelines.

## Session Info

```{r session-info}
sessionInfo()
```
