---
title: "Getting Started with Rwapor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Rwapor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The **Rwapor** package provides tools for downloading and processing WaPOR
(Water Productivity Open-access portal) and AgERA5 climate data from the
FAO GISMGR API. This vignette covers the basic workflow for:

1. Downloading raster maps for a region
2. Extracting time series with zonal statistics
3. Converting units between temporal resolutions

## Installation

```{r installation}
# Install from local source (development version)
install.packages("path/to/Rwapor", repos = NULL, type = "source")

# Load the package
library(Rwapor)
```

## Setting Up Parallel Processing

For faster downloads, set up parallel workers using the `future` package:

```{r setup-parallel}
library(future)
library(progressr)

# Use 4 parallel workers
future::plan(future::multisession, workers = 4)

# Enable progress reporting
progressr::handlers(global = TRUE)
```

## Available Variables

### WaPOR Variables

WaPOR provides water productivity and related variables at different spatial
and temporal resolutions:

```{r wapor-vars}
# List all WaPOR v3 variables
names(WAPOR3_VARS)

# Get metadata for a specific variable
get_variable_metadata("L1-AETI-D")
# $long_name: "Actual EvapoTranspiration and Interception"
# $units: "mm/day"
# $scale: 0.1
```

**Variable naming convention:** `{Level}-{Variable}-{TemporalResolution}`

- **Levels:** L1 (continental ~250m), L2 (regional ~100m), L3 (local)
- **Variables:** AETI, E, I, NPP, PCP, T, GBWP, NBWP
- **Temporal:** D (dekadal), M (monthly), A (annual), E (daily)

### AgERA5 Climate Variables

```{r agera5-vars}
# List AgERA5 variables
names(AGERA5_VARS)

# Example: Reference evapotranspiration
get_variable_metadata("AGERA5-ET0-E")
```

## Downloading Raster Maps

Use `wapor_map()` to download and save raster data for a region:

### Using a Bounding Box

```{r download-bbox}
# Define region as bounding box: c(xmin, ymin, xmax, ymax)
region <- c(35.0, 33.0, 36.0, 34.0)

# Download dekadal ET for January 2023
output_file <- wapor_map(
  region = region,
  variable = "L1-AETI-D",
  period = c("2023-01-01", "2023-01-31"),
  folder = "output"
)

print(output_file)
# "output/region_L1-AETI-D.tif"
```

### Using a Shapefile

```{r download-shapefile}
# Download for a shapefile region with local caching
output_file <- wapor_map(
  region = "path/to/my_region.shp",
  variable = "L2-NPP-M",
  period = c("2023-01-01", "2023-12-31"),
  folder = "output",
  filename = "npp_monthly_2023.tif",
  download_locally = TRUE  # Cache files locally first
)
```

### Using L3 Region Codes

For WaPOR Level 3 data, use the 3-letter region code:

```{r download-l3}
# Download for Awash Basin (Ethiopia)
output_file <- wapor_map(
  region = "AWA",
  variable = "L3-AETI-D",
  period = c("2023-06-01", "2023-06-30"),
  folder = "output"
)
```

## Extracting Time Series

Use `wapor_ts()` to extract zonal statistics (mean, min, max) for polygons:

### Basic Time Series Extraction

```{r ts-basic}
# Extract time series for a bounding box
df <- wapor_ts(
  region = c(35.0, 33.0, 36.0, 34.0),
  variable = "L1-AETI-D",
  period = c("2023-01-01", "2023-03-31")
)

head(df)
#   mean   min   max start_date   end_date  number_of_days layer_index
# 1  2.5   1.2   4.1 2023-01-01 2023-01-10             10           1
# 2  2.8   1.5   4.3 2023-01-11 2023-01-20             10           2
```

### Time Series for Multiple Polygons

```{r ts-polygons}
# Extract for multiple fields/polygons
df <- wapor_ts(
  region = "fields.geojson",
  variable = "L1-AETI-D",
  period = c("2023-01-01", "2023-12-31"),
  identifier = "field_name",  # Column to identify polygons
  download_locally = TRUE
)

# Results include statistics for each polygon at each time step
head(df)
```

### Accessing Metadata

The returned data frame includes metadata as attributes:

```{r ts-metadata}
# Get variable units
attr(df, "units")
# "mm/day"

# Get variable description
attr(df, "long_name")
# "Actual EvapoTranspiration and Interception"
```

## Unit Conversion

Convert values between temporal units using the `unit_conversion` parameter:

```{r unit-conversion}
# Convert dekadal data to monthly totals
df_monthly <- wapor_ts(
  region = c(35.0, 33.0, 36.0, 34.0),
  variable = "L1-AETI-D",
  period = c("2023-01-01", "2023-03-31"),
  unit_conversion = "month"
)

# Check the new units
attr(df_monthly, "units")
# "mm/month"

attr(df_monthly, "original_units")
# "mm/day"
```

Available conversions: `"none"`, `"day"`, `"dekad"`, `"month"`, `"year"`

## Working with URLs Directly

For advanced use cases, generate URLs without downloading:

```{r generate-urls}
# Get download URLs for a variable and period
urls <- wapor_generate_urls(
  variable = "L1-AETI-D",
  period = c("2023-01-01", "2023-01-31")
)

length(urls)
# 3 (one per dekad)

# Use with terra directly
library(terra)
r <- terra::rast(urls)
```

## Parsing Regions

The `parse_region()` function handles different region formats:
```{r parse-region}
# Bounding box
parse_region(c(35.0, 33.0, 36.0, 34.0))
# $type: "bbox"

# L3 region code
parse_region("AWA")
# $type: "l3_code"

# Vector file
parse_region("path/to/region.shp")
# $type: "vector"
```

## Error Handling

All functions provide informative error messages:

```{r error-handling, error=TRUE}
# Invalid variable format
wapor_generate_urls("INVALID")
# Error: Invalid variable format 'INVALID'. Expected format: 'LEVEL-VAR-TRES'

# Invalid date range
wapor_generate_urls("L1-AETI-D", period = c("2023-12-31", "2023-01-01"))
# Error: Start date must be before or equal to end date

# Invalid bounding box
parse_region(c(36, 33, 35, 34))
# Error: Invalid bounding box: xmin must be less than xmax
```

## Best Practices

1. **Use `download_locally = TRUE`** for large time series to avoid repeated
   network requests

2. **Set up parallel workers** before processing to speed up downloads:
   ```r
   future::plan(future::multisession, workers = 4)
   ```

3. **Check variable metadata** before downloading to understand units and
   scale factors:
   ```r
   get_variable_metadata("L1-AETI-D")
   ```

4. **Use appropriate temporal resolution** for your analysis:
   - Daily (E): Weather analysis
   - Dekadal (D): Crop monitoring
   - Monthly (M): Seasonal analysis
   - Annual (A): Long-term trends

## Session Info

```{r session-info}
sessionInfo()
```
