---
title: "Large Areas: Size-Aware and Tiled Processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Global Tiled Seasonal Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Rwapor decides how to process a job from its size. A 5 km farm at 20 m fits in
memory and runs without tiling; a 150 km irrigation scheme at 20 m, or several
years of dekads, is streamed or split into tiles automatically. The results are
the same whichever mode runs; only memory use and speed change.

## See the plan before you run

`wapor_plan_processing()` estimates the memory a job needs and reports the mode
it would use, with its reasons.

```{r}
library(Rwapor)

# A small farm at 20 m, one year of dekads
wapor_plan_processing(aoi = c(35.95, 33.80, 36.00, 33.85), resolution = 20, n_layers = 36)

# A large scheme at 20 m, four variables
wapor_plan_processing(aoi = c(35.0, 33.0, 36.5, 34.5), resolution = 20,
                      n_layers = 36, n_vars = 4)
```

The three modes are:

| Mode | What it does | When auto picks it |
|---|---|---|
| `memory` | Whole area and every dekad at once | Working set below 25% of the budget |
| `stream` | Whole area, dekads read in batches | Below the budget when streamed |
| `tiled` | Square tiles, assembled with a VRT | Otherwise, or when one layer is over 10% of the budget |

The budget is half the free RAM divided by the number of `future` workers. Set
it explicitly with `options(Rwapor.memory_budget_mb = 4096)`.

## Run a seasonal analysis

`wapor_run_seasonal_analysis()` plans and runs the job in one call. Leave
`processing` at `"auto"`, or force a mode.

```{r}
crop_mask <- wapor_load_crop_mask("scheme_crop_mask.tif")

config <- list(
  period = c("2023-01-01", "2023-12-31"),
  data_source = "api",
  aeti_var = "L3-AETI-D", ret_var = "L1-RET-D", precip_var = "L1-PCP-D",
  indicators = c("agg_aeti", "etc", "adequacy_etc", "agg_peff"),
  use_crop_mask = TRUE, use_season_rasters = FALSE,
  processing = "auto",          # or "memory", "stream", "tiled"
  output_dir = "scheme_run"     # stream and tiled outputs are written here
)
params <- wapor_create_crop_params(class_value = 1, crop_name = "Maize")

result <- wapor_run_seasonal_analysis(config, params, rasters = list(crop_mask = crop_mask))
result$processing          # the plan that ran
result$coverage$aeti       # share of season days with data, per pixel
```

When `aoi_region` is not given, the analysis area is the crop mask extent.

## Accuracy

Seasonal and monthly totals are summed at each source's own resolution and
resampled once onto the analysis grid. With the default nearest-neighbour
resampling, every value is a raw WaPOR value or a sum of raw values. For
example, AgERA5-based reference ET at about 9 km is summed on its own grid and
then assigned to the 20 m pixels inside each cell. To interpolate instead:

```{r}
config$resampling_method <- list(ret = "bilinear", pcp = "bilinear")
```

A pixel whose season is missing any dekad is `NA` by default rather than
treated as 0 mm. Allow partial seasons with, for example,
`config$min_coverage <- 0.9`.

## Parallel tiles

Tiles run in parallel under the active `future` plan. terra and GDAL caches are
divided between the workers automatically.

```{r}
future::plan(future::multisession, workers = 4)
result <- wapor_run_seasonal_analysis(config, params, rasters = list(crop_mask = crop_mask))
```

## Resumable tiled runs

`wapor_run_seasonal_analysis_tiled()` forces tiled mode and keeps a run
manifest with a checksum for every tile. If a run is interrupted, rerun it with
`resume = TRUE`: finished tiles are reused and only the missing ones are
computed.

```{r}
tiled <- wapor_run_seasonal_analysis_tiled(
  config, params, rasters = list(crop_mask = crop_mask),
  output_dir = "scheme_tiles", tile_size = 1024L, cog = TRUE, resume = TRUE
)
tiled$saved_files
```

## Check your setup

`inst/bench/remote_smoke_test.R` runs a short live check against the WaPOR API,
and `inst/bench/large_raster_benchmark.R` measures memory and time on synthetic
data:

```{r}
system.file("bench", "remote_smoke_test.R", package = "Rwapor")
```
