---
title: "Using the Rwapor Shiny Dashboard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Rwapor Shiny Dashboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The Rwapor package includes an interactive Shiny dashboard that provides a user-friendly interface for downloading and analyzing WaPOR and AgERA5 data. This vignette covers the dashboard's features, with special focus on the **L3 Region Auto-Detection** system.

## Launching the Dashboard

```{r launch}
library(Rwapor)

# Start the dashboard
run_wapor()
```

The dashboard will open in your default web browser.

## Dashboard Overview

The dashboard consists of three main tabs:

1. **Download** - Download WaPOR and AgERA5 raster data
2. **Analysis** - Perform crop water productivity analysis
3. **Seasonal Comparison** - Compare seasonal patterns across multiple years

## Download Tab Workflow

The Download tab follows an intuitive step-by-step workflow:

### 1. Project Folder

Specify the output directory where downloaded files will be saved.

- **Quick Access**: Use the folder icon to browse for a directory
- **Favorites**: Star frequently-used folders for quick access
- **Default**: `wapor_project` in your current working directory

### 2. Area of Interest (AOI)

Define your study area before selecting variables. This enables automatic L3 region detection.

#### AOI Selection Methods

**A. Draw on Map**
- **Rectangle**: Click twice to define opposite corners
- **Polygon**: Click multiple points, then click "Finish"
- Use "Clear" to reset

**B. Upload New**
- Supported formats:
  - **Vector**: Shapefile (.shp), GeoJSON, GeoPackage, KML
  - **Raster**: GeoTIFF (.tif/.tiff), NetCDF (.nc), ESRI Grid (.grd)
- For rasters: the extent is automatically extracted
- Star files for quick access in future sessions

**C. Project Assets**
- Scan the project folder for existing spatial files
- Quickly reuse AOIs from previous analyses

### 3. Variables

Select one or more WaPOR or AgERA5 variables to download.

**Variable Levels:**
- **L1**: Continental scale (~250m resolution)
- **L2**: Country/regional scale (~100m resolution)
- **L3**: Sub-national / irrigation scheme scale (~10-30m resolution)

**Temporal Resolutions:**
- **A**: Annual
- **M**: Monthly
- **D**: Dekadal (10-day periods)
- **E**: Daily

### 4. L3 Region Selection (for L3 variables)

When L3 variables are selected, the L3 Region dropdown appears.

---

## L3 Region Auto-Detection Feature

**New in v0.9.3**: The dashboard automatically detects which L3 regions overlap with your selected AOI.

### How It Works

When you define an AOI (via drawing, upload, or project assets), the system:

1. **Detects Spatial Overlap**: Uses `wapor_guess_region()` to identify L3 regions that intersect with your AOI
2. **Filters Dropdown**: Shows only overlapping regions in the L3 Region dropdown
3. **Auto-Selects**: If exactly one region overlaps, it's automatically selected
4. **Provides Feedback**: Displays status messages to guide your selection

### Status Messages

The dashboard provides real-time feedback about L3 region detection:

#### ✅ Success (Green)
**"Auto-selected: Ethiopia - Awash"**
- Exactly one L3 region overlaps your AOI
- The region is automatically selected
- You can proceed with download

#### ⚠️ Multiple Overlaps (Yellow)
**"Multiple L3 regions overlap (AWA, KOG, BUS). Please select one."**
- Your AOI spans multiple L3 regions
- The dropdown is filtered to show only overlapping regions
- You must manually select which region to download

#### ⚠️ No Overlap (Yellow)
**"No L3 regions overlap with this AOI"**
- Your AOI doesn't intersect any L3 regions
- L3 data is not available for this area
- Consider using L1 or L2 variables instead

#### ℹ️ Info (Blue)
**"Select an AOI to auto-detect L3 regions"**
- No AOI has been defined yet
- All L3 regions remain available in the dropdown
- Define an AOI to enable auto-detection

### Use Cases

#### Use Case 1: Single Irrigation Scheme

You want to download L3 data for the Awash Basin in Ethiopia:

```r
# 1. Launch dashboard
run_wapor()

# 2. In the browser:
#    - Set Project Folder: "~/my_projects/awash_analysis"
#    - Upload AOI: Upload a shapefile of Awash Basin boundary
#    - Select Variable: L3-AETI-D (dekadal evapotranspiration)
#    - Result: "Auto-selected: Ethiopia - Awash (AWA)"
#    - Select Time Period: 2023-01-01 to 2023-12-31
#    - Click "Download Data"
```

#### Use Case 2: Regional Study with Multiple Schemes

You're analyzing water productivity across East Africa:

```r
# 1. Draw a large rectangle covering Kenya
# 2. Select L3-GBWP-A (gross biomass water productivity)
# 3. Result: "Multiple L3 regions overlap (BUS, KMW, KTB)"
# 4. Manually select one region (e.g., KMW - Mwea)
# 5. Download completes for selected region
```

#### Use Case 3: Raster AOI

You have a crop classification raster and want matching L3 data:

```r
# 1. Upload your raster file (e.g., "crop_mask.tif")
# 2. System extracts raster extent automatically
# 3. Detects overlapping L3 regions based on extent
# 4. Auto-selects if extent overlaps exactly one region
# 5. Download proceeds with spatial alignment
```

### Technical Implementation

The auto-detection uses these package functions under the hood:

```{r technical}
# Parse AOI (bbox or file path)
region_info <- wapor_parse_region(aoi_bbox)

# Detect overlapping L3 regions
overlapping_codes <- wapor_guess_region(
  variable = "L3-AETI-D",
  reg_info = region_info,
  period = c("2023-01-01", "2023-01-11")  # Sample period for detection
)

# Result: character vector of L3 codes, e.g., c("AWA", "KOG")
```

**Key Functions:**
- `wapor_parse_region()`: Standardizes AOI input (vector, bbox, or raster)
- `wapor_guess_region()`: Performs spatial intersection analysis
- `wapor_l3_extent()`: Retrieves cached L3 region boundaries

### Performance Notes

- **First Detection**: May take 5-10 seconds as L3 extents are downloaded and cached
- **Subsequent Detections**: Nearly instant using disk cache
- **Cache Location**: `{R_user_dir}/Rwapor/cache/l3_extents/`

The cache persists across R sessions for fast re-detection.

---

## Time Period

Select the date range for data download:

- **Start Date**: First day of the period
- **End Date**: Last day of the period
- **Temporal Coverage**: Depends on variable (WaPOR: 2009-present, AgERA5: 1979-present)

---

## Output Settings

### File Options

**Seasonal Aggregate**
- Calculates sum (flux variables) or mean (intensive properties) over the entire period
- Produces one file per variable for the full time range

**Save Per-Timestep Files**
- Saves individual files for each dekad/day/month
- Useful for time series analysis

**Both Options**
- Check both to get individual files + one seasonal aggregate
- Recommended for comprehensive analysis

### Unit Conversion

The Download tab now exposes only the two safe public modes:

- **Match variable temporal unit**: use variable-specific behavior
  - dekadal daily-rate products are converted to dekadal totals
  - monthly products stay monthly
  - annual products stay annual
- **No conversion (raw API values)**: keep the original API values without temporal conversion

This avoids mixed-download mistakes such as forcing monthly products into
dekadal scaling.

---

## Download Execution

Click **"Download Data"** to start the download process.

### Progress Tracking

- **Progress bar**: Shows overall download progress
- **Status messages**: Indicate current step (e.g., "Stage 1 of 2: Individual files for L1-AETI-D...")
- **Completion notification**: Displays number of files downloaded

### Error Handling

Common issues and solutions:

**"No L3 regions overlap with this AOI"**
- Solution: Use L1 or L2 variables, or select a different AOI

**"Multiple L3 regions overlap. Please select one."**
- Solution: Manually choose one region from the filtered dropdown

**"Region cannot be NULL"**
- Solution: Define an AOI before downloading

**Download stalls or fails**
- Check internet connection
- Verify date range is within data availability
- Try smaller spatial/temporal extent

---

## R Code Preview

The dashboard generates equivalent R code for reproducibility. Click the **"R Code Preview"** button at the bottom of the map to view and copy the code.

Example generated code:

```{r example-code}
library(Rwapor)

region   <- c(35.5, 8.0, 40.5, 12.0)  # bbox
period   <- c("2023-01-01", "2023-12-31")
variable <- "L3-AETI-D"
folder   <- "~/wapor_project"

map_paths <- wapor_map(
  region = region,
  variable = variable,
  period = period,
  folder = folder,
  unit_conversion = "unit_conversion",
  seasonal = FALSE,
  separate_files = TRUE,
  mask = FALSE
)
```

You can copy this code to run downloads programmatically without the dashboard.

---

## Tips and Best Practices

### Efficient AOI Definition

1. **Raster AOI for Exact Matching**: Upload your analysis raster (e.g., crop mask) to ensure perfect spatial alignment
2. **Favorites for Repeated Analyses**: Star frequently-used shapefiles and folders
3. **Project Assets**: Keep all AOI files in your project folder for easy scanning

### L3 Region Selection Strategy

1. **Define AOI First**: Always set your AOI before selecting L3 variables to leverage auto-detection
2. **Use Specific Boundaries**: For multi-region studies, use precise boundaries to avoid ambiguous overlaps
3. **Check Coverage**: If auto-detection shows no overlap, verify L3 data exists for your area using the L3_REGIONS list:

```{r check-coverage}
# List all available L3 regions
names(L3_REGIONS)

# Find regions in a specific country
Filter(function(r) r$country == "Kenya", L3_REGIONS)
```

### Download Optimization

1. **Spatial Extent**: Smaller areas download faster
2. **Temporal Range**: Limit date ranges to reduce file count
3. **Variable Selection**: Download multiple variables in one operation for efficiency
4. **Seasonal Aggregate**: Use seasonal option to reduce storage when time series isn't needed

### Debugging Downloads

If downloads fail or produce unexpected results:

1. **Check Generated Code**: Review the R Code Preview for correctness
2. **Test Programmatically**: Copy the generated code and run with `verbose = TRUE`
3. **Verify Metadata**: Use `wapor_fetch_metadata()` to confirm variable availability
4. **Spatial Validation**: Use `wapor_compare_geom()` to check AOI alignment

---

## Integration with Programmatic Workflow

You can seamlessly combine dashboard and programmatic approaches:

```{r integration}
# 1. Use dashboard to:
#    - Explore available variables
#    - Test AOI definitions
#    - Verify L3 region detection
#    - Generate initial download code

# 2. Copy generated code to your script
# 3. Extend with programmatic analysis:

library(Rwapor)
library(terra)

# Download using dashboard-generated parameters
aeti_files <- wapor_map(
  region = "AWA",  # Auto-detected L3 code
  variable = "L3-AETI-D",
  period = c("2023-01-01", "2023-12-31"),
  folder = "~/wapor_project"
)

# Continue with analysis
aeti_stack <- rast(aeti_files)
seasonal_mean <- mean(aeti_stack)
plot(seasonal_mean, main = "Mean Dekadal AETI (2023)")
```

---

## Keyboard Shortcuts

When drawing polygons on the map:

- **Escape**: Cancel drawing
- **Backspace**: Remove last point
- **Enter**: Finish polygon (alternative to "Finish" button)

---

## Troubleshooting

### L3 Auto-Detection Not Working

**Symptoms**: No status message appears, or dropdown doesn't filter

**Solutions**:
1. Ensure L3 variables are selected
2. Verify AOI is defined (check "Selected ROI" display)
3. Check date range is valid
4. Clear browser cache and reload dashboard

### "Multiple regions overlap" But I Expected One

**Cause**: AOI extends slightly into adjacent L3 regions

**Solution**:
- Refine AOI boundary to be fully within one region
- Or manually select the primary region of interest

### Raster AOI Not Detected

**Symptoms**: Uploaded raster but no overlap detected

**Solutions**:
1. Verify raster has valid CRS metadata (`gdalinfo file.tif`)
2. Check raster extent overlaps expected L3 region
3. Ensure file format is supported (.tif, .tiff, .nc, .grd)

---

## Advanced: Extending the Dashboard

The dashboard is built with modular Shiny components. Key modules:

- `mod_download_ui/server`: Download workflow and L3 detection
- `mod_aoi_ui/server`: AOI selection and spatial operations
- `mod_analysis_ui/server`: Crop water productivity analysis

To extend functionality, modify these modules in `inst/shiny/`.

---

## Further Reading

- **Getting Started**: `vignette("getting-started", package = "Rwapor")`
- **Data Catalog**: `vignette("data-catalog", package = "Rwapor")`
- **Function Reference**: `help(package = "Rwapor")`

---

## Summary

The Rwapor Shiny dashboard provides an intuitive interface for WaPOR/AgERA5 data access, with the L3 region auto-detection feature significantly streamlining the workflow for sub-national analyses. By defining your AOI first, the system automatically identifies and filters relevant L3 regions, reducing errors and improving efficiency.

Key Takeaways:

1. ✅ Define AOI before selecting L3 variables
2. ✅ Raster AOIs work seamlessly for spatial alignment
3. ✅ Auto-detection handles single-region cases automatically
4. ✅ Multi-region overlaps require manual selection
5. ✅ Use R Code Preview for reproducible workflows
