---
title: "Water Productivity Analysis with Rwapor"
subtitle: "Training Notebook — Worked Examples for Citrus and Wheat"
author: "Rwapor Training Series"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    df-print: paged
    theme: flatly
    fig-width: 8
    fig-height: 5.5
execute:
  warning: false
  message: false
---

# About This Notebook

This is a **standalone, independent Quarto notebook** — it is not part of the `Rwapor`
package build (vignettes, `R CMD check`, and pkgdown never see it). It calls only
**exported** `Rwapor` functions; nothing in the package itself is modified. Its purpose
is training: every step below downloads or computes something real, shows you the
result as a plot, map, or summary table, and explains in plain language — with the
underlying equation — what just happened.

Two worked examples run end-to-end:

| | Part A — Citrus | Part B — Wheat |
|---|---|---|
| Input | A single polygon layer used as **both** the mask and the AOI | A **cereal mask raster** (may hold several crop classes) plus a **separate AOI boundary** polygon |
| Season | 2024-03-01 → 2025-02-28 (one year, spans the calendar year-end) | 2023-11-01 → 2024-05-31 (a single winter-wheat season) |
| Crop parameters | Built from scratch (`Rwapor` ships no built-in Citrus Kc profile — see §2.2) | Reused from `Rwapor`'s built-in "Winter Wheat" FAO-56 profile |
| Indicators | Full chain: AETI, T, RET, PCP, NPP, ETc, both Adequacy ratios, Beneficial Fraction, Green/Blue water, Yield, Biomass, CWP, BWP | Same full chain |

::: {.callout-important}
## Before you run this

Every file path below (`citrus_fields.geojson`, `cereal_mask.tif`, `wheat_aoi.geojson`,
…) is a **placeholder**. Replace the values in the "Point to your input files" chunk at
the start of each Part with your own data before running. Everything downstream reads
those variables — nothing else in the notebook needs to change.

You'll also need a working internet connection: both examples stream data live from the
FAO WaPOR v3 API (`data_source = "api"`), so every plot and table you see is computed
from real, freshly-downloaded rasters, not mock data. No API key is required.
:::

::: {.callout-note collapse="true"}
## Two things I verified against the package source while building this notebook

**1. `aoi_region` matters for API runs.** `wapor_run_seasonal_analysis()` takes an
optional `aoi_region` argument that scopes *every* raster it streams to your AOI before
loading it into memory. If you omit it, the function still runs — but it downloads the
**full extent** of whatever WaPOR tile covers your variable/period (e.g. an entire L2
tile) before your `crop_mask` ever gets a chance to narrow it down, which is slow and
memory-heavy. This notebook always passes `aoi_region` explicitly.

**2. The indicator code for effective precipitation is `"agg_peff"`, not `"peff"`.**
(`R/analysis_engine.R` checks for the string `"agg_peff"`.) This notebook uses the
working code throughout.
:::

---

# 1. Installation and Setup

## 1.1 Install `Rwapor` and Supporting Packages

```{r}
#| label: install
#| eval: false

# install.packages("remotes")
remotes::install_github("almutaz9000/Rwapor")

# Supporting packages used only by this training notebook (all already
# Suggested by Rwapor itself, except sf/dplyr/ggplot2 which are Imports)
install.packages(c("terra", "sf", "dplyr", "ggplot2", "leaflet", "patchwork"))
```

## 1.2 Load Packages

```{r}
#| label: setup

library(Rwapor)
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(leaflet)
library(patchwork)
```

`Rwapor` tunes GDAL's `/vsicurl/` streaming settings automatically on load (larger
chunk reads instead of GDAL's tiny default), which is what makes streaming individual
WaPOR dekads directly from the API practical instead of downloading everything first.

---

# 2. Part A — Citrus Water Productivity

## 2.1 Point to Your Input Files

```{r}
#| label: citrus-config

# --- EDIT THESE THREE LINES FOR YOUR OWN DATA -------------------------------
citrus_polygon_path <- "data/citrus/citrus_fields.geojson"  # used as BOTH mask and AOI
citrus_period        <- c("2024-03-01", "2025-02-28")        # 1 Mar -> end of Feb (next yr)
citrus_download_dir  <- "wapor_data/citrus"
# -----------------------------------------------------------------------------

citrus_output_dir <- "outputs/Citrus_2024_2025"
```

::: {.callout-note}
## Why one file for both mask and AOI?

For Citrus you told us the same polygon layer plays two roles: it defines **where** to
download data (the AOI) and **which pixels count as citrus** (the mask). `wapor_map()`
and `wapor_run_seasonal_analysis()` both accept a vector file directly for this, so
`citrus_polygon_path` is reused unchanged in every step below — no rasterizing is needed
until §2.4, and even then it's rasterizing the *same* file.
:::

## 2.2 Explore the Citrus AOI on an Interactive Map

```{r}
#| label: citrus-explore-aoi
#| fig-cap: "Citrus field boundaries used as the AOI/mask"

citrus_poly <- sf::st_read(citrus_polygon_path, quiet = TRUE)
citrus_poly <- sf::st_transform(citrus_poly, 4326)  # leaflet expects WGS84

leaflet(citrus_poly) |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addPolygons(
    color = "#e6b800", weight = 2, fillOpacity = 0.15,
    label = ~as.character(seq_len(nrow(citrus_poly)))
  ) |>
  addLegend(colors = "#e6b800", labels = "Citrus field boundary", position = "bottomright")
```

```{r}
#| label: citrus-aoi-summary

data.frame(
  n_polygons  = nrow(citrus_poly),
  crs         = sf::st_crs(citrus_poly)$input,
  area_ha     = round(sum(as.numeric(sf::st_area(citrus_poly))) / 10000, 1)
)
```

## 2.3 Define Citrus Crop Water-Use Parameters

::: {.callout-note}
## Why we can't just reuse an FAO default here

`Rwapor`'s built-in `FAO_CROP_DEFAULTS` table (`wapor_list_crops()`) ships profiles for
Winter Wheat, Maize, Rice, Cotton, Potato, and eight other annual/perennial field crops
— but **not Citrus**. `wapor_custom_crop(base_crop = "Citrus", ...)` would fail with
*"Base crop not found"*. So instead we build the profile from scratch with
`wapor_create_crop_params()`, sourcing real numbers rather than guessing:

- **Kc values** — FAO Irrigation & Drainage Paper 56, Table 6.2, "Citrus, 70% canopy, no
  ground cover": $K_{c,ini}=0.75$, $K_{c,mid}=0.70$, $K_{c,end}=0.75$.
- **Growth-stage lengths** — this repo's own `fao_growth_stages.csv` already has a
  Citrus row (Mediterranean region): $L_{ini}=60$, $L_{dev}=90$, $L_{mid}=120$,
  $L_{late}=95$ days (365-day total). `Rwapor`'s engine only needs $L_{ini}$, $L_{mid}$,
  $L_{late}$ — it derives $L_{dev}$ itself as whatever remains of the *actual* season
  length once those three are subtracted, so it self-adjusts to your real season.
:::

```{r}
#| label: citrus-crop-params

citrus_params <- wapor_create_crop_params(
  class_value  = 1L,                # matches the value we'll rasterize the mask to
  crop_name    = "Citrus (70% canopy, no ground cover)",
  kc_ini       = 0.75,
  kc_mid       = 0.70,
  kc_end       = 0.75,
  l_ini_days   = 60L,
  l_mid_days   = 120L,
  l_late_days  = 95L,
  max_height_m = 3.0,                # typical mature citrus tree height
  HI           = 0.55,               # Harvest Index: fruit mass / total tree biomass increment
  MC           = 0.85,               # Moisture Content of fresh fruit at harvest
  fc           = 0.70,               # ground-cover fraction used in the Kc table above
  AOT          = 0.8,                # above-ground / total biomass ratio (perennial default)
  region       = "Mediterranean",
  notes        = "FAO-56 Table 6.2 (Kc) + fao_growth_stages.csv (stage lengths); no built-in FAO_CROP_DEFAULTS profile exists for Citrus"
)

citrus_params
```

### The Crop Coefficient (Kc) Curve

`Rwapor` builds a **four-stage daily Kc curve** — flat at $K_{c,ini}$, linear ramp up
through the development stage, flat at $K_{c,mid}$, linear ramp down to $K_{c,end}$ —
and this is what drives crop water demand ($ET_c$, §2.6) all season long:

```{r}
#| label: citrus-kc-curve
#| fig-cap: "Daily Kc curve for Citrus over a 365-day season"

citrus_l_dev <- 365L - (citrus_params$l_ini_days + citrus_params$l_mid_days + citrus_params$l_late_days)

citrus_kc_daily <- wapor_build_kc(
  kc_ini = citrus_params$kc_ini, kc_mid = citrus_params$kc_mid, kc_end = citrus_params$kc_end,
  l_ini  = citrus_params$l_ini_days, l_dev = citrus_l_dev,
  l_mid  = citrus_params$l_mid_days, l_late = citrus_params$l_late_days
)

wapor_plot_kc_curve(citrus_kc_daily) +
  ggplot2::labs(title = "Citrus Kc Curve", subtitle = "FAO-56 four-stage model")
```

## 2.4 Obtain the Seasonal AETI Raster Map

This is the step you specifically asked for: download WaPOR's dekadal Actual
EvapoTranspiration and Interception ($AETI$, mm/day) for the citrus AOI and season, and
let `wapor_map()` aggregate every overlapping dekad into a **single seasonal total**
(mm over the whole season).

::: {.callout-note}
## What "seasonal aggregation" actually does

$AETI$ is a *flux* variable (mm/**day**) — `Rwapor` knows this from its variable
metadata and therefore **weighted-sums** it, not averages it: each dekadal raster is
multiplied by however many days of the season it overlaps, then all dekads are summed.
A 10-day dekad reading 2.5 mm/day contributes $2.5 \times 10 = 25$ mm to the seasonal
total. This is the same "aggregation rule" logic that decides whether any WaPOR/AgERA5
variable should be summed (fluxes like AETI, NPP, PCP) or averaged (state variables
like RSM, temperature).
:::

```{r}
#| label: citrus-seasonal-aeti-map
#| fig-cap: "Seasonal AETI (mm) for the citrus AOI, Mar 2024 -- Feb 2025"

citrus_aeti_seasonal_path <- wapor_map(
  region    = citrus_polygon_path,
  variable  = "L2-AETI-D",
  period    = citrus_period,
  folder    = citrus_download_dir,
  seasonal  = TRUE,
  mask      = TRUE
)

citrus_aeti_r <- terra::rast(citrus_aeti_seasonal_path)
wapor_plot_map(citrus_aeti_r, indicator = "Seasonal AETI (mm)", title = "Citrus — Seasonal AETI")
```

```{r}
#| label: citrus-seasonal-aeti-leaflet
#| fig-cap: "Interactive view of the seasonal AETI raster over the citrus AOI"

aeti_pal <- colorNumeric("YlGnBu", terra::values(citrus_aeti_r), na.color = "transparent")

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(citrus_aeti_r, colors = aeti_pal, opacity = 0.85) |>
  addPolygons(data = citrus_poly, fill = FALSE, color = "black", weight = 1) |>
  addLegend(pal = aeti_pal, values = terra::values(citrus_aeti_r), title = "AETI (mm)")
```

## 2.5 Rasterize the Citrus Mask

We already have a WaPOR-grid raster from the step above (`citrus_aeti_r`) — it makes a
ready-made template (same resolution, extent, and CRS) to rasterize the citrus polygons
onto:

```{r}
#| label: citrus-rasterize-mask
#| fig-cap: "Citrus crop mask (class 1) on the WaPOR grid"

citrus_vect <- terra::vect(citrus_poly)
citrus_mask <- terra::rasterize(citrus_vect, citrus_aeti_r, field = 1, background = NA)
names(citrus_mask) <- "crop_mask"

wapor_plot_map(citrus_mask, indicator = "Citrus mask (class 1)", title = "Citrus Crop Mask")
```

## 2.6 Run the Full Seasonal Water Productivity Analysis

### The Indicators and Their Equations

| Indicator | Code | Equation |
|---|---|---|
| Seasonal AETI, T, RET, PCP, NPP | `agg_aeti`, `agg_t`, `agg_ret`, `agg_pcp`, `agg_npp` | Weighted seasonal sum of each dekadal flux (see §2.4 callout) |
| Crop Evapotranspiration | `etc` | $ET_c(t) = K_c(t) \times RET(t)$, summed over the season using the Kc curve built in §2.3 |
| ETc-based Adequacy | `adequacy_etc` | $Adequacy_{ETc} = AETI_{seasonal} / ET_{c,seasonal}$ — a ratio near 1 means actual water consumption matched crop demand; well below 1 flags likely water stress |
| P95-based Adequacy | `adequacy_p95` | $Adequacy_{P95} = AETI_{seasonal} / P_{95}(AETI_{seasonal})$ within the crop mask — compares every pixel to the *best-performing 5%* of the same field(s), independent of any Kc assumption |
| Beneficial Fraction | `beneficial_fraction` | $BF = T_{seasonal} / AETI_{seasonal}$ — the share of consumed water that went to productive plant transpiration ($T$) rather than non-productive soil evaporation/interception |
| Effective Precipitation | `agg_peff` | USDA-SCS method: $P_{eff} = P\cdot\frac{125 - 0.2P}{125}$ if $P \le 250$ mm/month, else $P_{eff} = 125 + 0.1P$ |
| Green Water | `green_water` | $\min(AETI_{seasonal}, P_{eff,seasonal})$ — the part of consumption met by rainfall |
| Blue Water | `blue_water` | $\max(0,\ AETI_{seasonal} - P_{eff,seasonal})$ — the part met by irrigation/other blue-water sources |
| Biomass (Dry Matter Production) | `agg_biomass_kg`/`agg_biomass_t` | $DMP = NPP_{gC/m^2} \times 22.222$ (WaPOR's standard gC → dry-matter conversion) |
| Yield | `yield_npp` | $AGBM = \dfrac{AOT \times f_c \times DMP / (1-MC)}{1000}$ (t/ha), then $Yield = HI \times AGBM$ |
| Crop Water Productivity | `cwp_bwp` (CWP) | $CWP = \dfrac{Yield_{kg/ha}}{AETI_{mm} \times 10}$ kg/m³ (the ×10 converts mm over 1 ha into m³) — productivity of the **harvestable** product |
| Biomass Water Productivity | `cwp_bwp` (BWP) | Same formula, using total $DMP$ instead of harvestable $Yield$ — productivity of **all** biomass grown |

```{r}
#| label: citrus-all-indicators

all_indicators <- c(
  "agg_aeti", "agg_t", "agg_ret", "agg_pcp", "agg_npp", "agg_peff",
  "etc", "adequacy_etc", "adequacy_p95", "beneficial_fraction",
  "green_water", "blue_water", "cwp_bwp", "yield_npp",
  "agg_biomass_kg", "agg_biomass_t"
)
```

```{r}
#| label: citrus-run-analysis

citrus_config <- list(
  period        = citrus_period,
  ref_year      = 1970,             # internal day-of-year baseline; leave as-is
  aeti_var      = "L2-AETI-D",      # or "L1-AETI-D" (~300 m, global) / "L3-AETI-D" (~20 m, L3 schemes only)
  ret_var       = "L1-RET-D",       # RET is only published at Level 1 (~5 km, reanalysis-based)
  precip_var    = "L1-PCP-D",       # PCP is only published at Level 1 (~5 km, gridded precipitation)
  npp_var       = "L2-NPP-D",       # or "L1-NPP-D" (~300 m, global)
  t_var         = "L2-T-D",         # or "L1-T-D" (~300 m, global)
  data_source   = "api",            # or "local" to read files already downloaded into `folder`
  folder        = citrus_download_dir,
  area_weighted = TRUE,             # corrects for latitude-dependent pixel-area distortion (EPSG:4326)
  indicators    = all_indicators
)

citrus_season <- wapor_run_seasonal_analysis(
  config      = citrus_config,
  crop_params = citrus_params,
  rasters     = list(crop_mask = citrus_mask),
  aoi_region  = citrus_polygon_path   # keeps every downloaded raster scoped to the AOI (see intro callout)
)

names(citrus_season)
```

## 2.7 Explore the Outputs, Step by Step

### Seasonal AETI and Transpiration

```{r}
#| label: citrus-out-aeti-t
#| fig-cap: "Seasonal AETI and Transpiration (T) for the citrus fields"

p1 <- wapor_plot_map(citrus_season$seasonal_aeti$raster, indicator = "AETI (mm)", title = "Seasonal AETI")
p2 <- wapor_plot_map(citrus_season$seasonal_t$raster, indicator = "T (mm)", title = "Seasonal Transpiration")
patchwork::wrap_plots(p1, p2)

citrus_season$seasonal_aeti_by_class
```

### Crop Water Demand (ETc) and Adequacy

```{r}
#| label: citrus-out-etc-adequacy
#| fig-cap: "Water adequacy: AETI vs. crop water demand"

citrus_etc <- citrus_season$etc_by_class[["1"]]$etc_seasonal
wapor_plot_map(citrus_etc, indicator = "ETc (mm)", title = "Seasonal Crop Water Demand (ETc)")

wapor_plot_map(citrus_season$adequacy_etc, indicator = "Adequacy", title = "Water Adequacy (AETI / ETc)")

data.frame(
  mean_etc_mm          = round(terra::global(citrus_etc, "mean", na.rm = TRUE)[1, 1], 1),
  mean_adequacy_etc    = round(terra::global(citrus_season$adequacy_etc, "mean", na.rm = TRUE)[1, 1], 2),
  mean_adequacy_p95    = round(terra::global(citrus_season$adequacy_p95, "mean", na.rm = TRUE)[1, 1], 2)
)
```

### Beneficial Fraction

```{r}
#| label: citrus-out-bf
#| fig-cap: "Fraction of consumed water that was productive transpiration"

wapor_plot_map(citrus_season$beneficial_fraction, indicator = "T/AETI", title = "Beneficial Fraction")
```

A beneficial fraction well below ~0.6–0.7 for a mature orchard usually points to a lot
of water being lost to soil evaporation or interception rather than going through the
trees themselves — worth cross-checking with an irrigation/mulching survey on the
ground.

### Water Source: Green vs. Blue Water

```{r}
#| label: citrus-out-green-blue
#| fig-cap: "Rainfall-sourced (green) vs. irrigation-sourced (blue) water consumption"

p3 <- wapor_plot_map(citrus_season$green_water, indicator = "Green water (mm)", title = "Green Water")
p4 <- wapor_plot_map(citrus_season$blue_water, indicator = "Blue water (mm)", title = "Blue Water")
patchwork::wrap_plots(p3, p4)

data.frame(
  mean_green_mm = round(terra::global(citrus_season$green_water, "mean", na.rm = TRUE)[1, 1], 1),
  mean_blue_mm  = round(terra::global(citrus_season$blue_water, "mean", na.rm = TRUE)[1, 1], 1)
)
```

### Yield, Biomass, and Water Productivity

```{r}
#| label: citrus-out-productivity
#| fig-cap: "Citrus yield raster (t/ha)"

wapor_plot_map(citrus_season$yield_raster, indicator = "Yield (t/ha)", title = "Estimated Citrus Yield")

data.frame(
  indicator = c("CWP (kg/m3)", "BWP (kg/m3)"),
  value     = round(c(citrus_season$cwp, citrus_season$bwp), 3)
)
```

$CWP$ tells you how much harvestable fruit you got per m³ of water consumed; $BWP$
tells you the same thing for *all* biomass the tree grew (wood, leaves, and fruit
together) — comparing the two is a quick check on how much of the tree's total growth
ended up as marketable product.

## 2.8 Export All Outputs

```{r}
#| label: citrus-export

wapor_export_analysis_outputs(
  results      = citrus_season,
  folder       = citrus_output_dir,
  indicators   = all_indicators,
  season_label = "Citrus_2024_2025"
)
```

---

# 3. Part B — Wheat Water Productivity

## 3.1 Point to Your Input Files

```{r}
#| label: wheat-config

# --- EDIT THESE FOR YOUR OWN DATA --------------------------------------------
cereal_mask_path <- "data/wheat/cereal_mask.tif"     # a raster; may hold several crop classes
wheat_aoi_path    <- "data/wheat/wheat_aoi.geojson"   # a SEPARATE study-area boundary
wheat_class_code  <- 1L                               # the class code for wheat in cereal_mask_path
wheat_period      <- c("2023-11-01", "2024-05-31")
wheat_download_dir <- "wapor_data/wheat"
# -------------------------------------------------------------------------------

wheat_output_dir <- "outputs/Wheat_Winter2023_2024"
```

::: {.callout-note}
## Why Wheat is a different shape of input than Citrus

You told us you already have a **cereal mask raster** (possibly covering several
cereal crops, each with its own class code) and a **separate AOI boundary**. That's a
more common real-world setup than Citrus's single polygon: the AOI defines the download
extent, and the mask raster — after we isolate wheat's class code — decides which
pixels the seasonal engine treats as wheat.
:::

## 3.2 Explore the AOI and the Cereal Mask

```{r}
#| label: wheat-explore-inputs
#| fig-cap: "Study-area AOI and the raw cereal mask (all classes)"

wheat_aoi <- sf::st_read(wheat_aoi_path, quiet = TRUE) |> sf::st_transform(4326)
cereal_mask_raw <- terra::rast(cereal_mask_path)

# Which class codes actually exist in this mask, and how many pixels each has:
terra::freq(cereal_mask_raw)
```

```{r}
#| label: wheat-leaflet-inputs
#| fig-cap: "Interactive view: AOI boundary over the raw cereal mask"

class_pal <- colorNumeric("Set2", terra::values(cereal_mask_raw), na.color = "transparent")

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(cereal_mask_raw, colors = class_pal, opacity = 0.7) |>
  addPolygons(data = wheat_aoi, fill = FALSE, color = "black", weight = 2) |>
  addLegend(pal = class_pal, values = terra::values(cereal_mask_raw), title = "Cereal class code")
```

Use the frequency table above to confirm `wheat_class_code` in §3.1 actually matches
wheat's legend entry before continuing.

## 3.3 Define Wheat's Crop Water-Use Parameters

Unlike Citrus, `Rwapor` does ship a built-in **Winter Wheat** profile
(`wapor_list_crops()`), so we reuse it directly instead of building one from scratch:

```{r}
#| label: wheat-crop-params

wheat_params <- wapor_custom_crop(
  base_crop   = "Winter Wheat",
  class_value = wheat_class_code,   # must match the isolated mask below
  crop_name   = "Irrigated Winter Wheat",
  kc_ini      = 0.35,
  kc_mid      = 1.15,
  kc_end      = 0.30,
  HI          = 0.45,               # Harvest Index (grain mass / total biomass)
  MC          = 0.13                # Moisture Content of grain at harvest
)

wheat_params
```

```{r}
#| label: wheat-kc-curve
#| fig-cap: "Daily Kc curve for Winter Wheat over the 2023-11-01 to 2024-05-31 season"

wheat_total_days <- as.integer(as.Date(wheat_period[2]) - as.Date(wheat_period[1])) + 1L
wheat_l_dev <- wheat_total_days - (wheat_params$l_ini_days + wheat_params$l_mid_days + wheat_params$l_late_days)

wheat_kc_daily <- wapor_build_kc(
  kc_ini = wheat_params$kc_ini, kc_mid = wheat_params$kc_mid, kc_end = wheat_params$kc_end,
  l_ini  = wheat_params$l_ini_days, l_dev = wheat_l_dev,
  l_mid  = wheat_params$l_mid_days, l_late = wheat_params$l_late_days
)

wapor_plot_kc_curve(wheat_kc_daily) +
  ggplot2::labs(title = "Winter Wheat Kc Curve", subtitle = "FAO-56 four-stage model, fitted to a 212-day season")
```

## 3.4 Obtain the Seasonal AETI Raster Map

```{r}
#| label: wheat-seasonal-aeti-map
#| fig-cap: "Seasonal AETI (mm) over the wheat study area, Nov 2023 -- May 2024"

wheat_aeti_seasonal_path <- wapor_map(
  region    = wheat_aoi_path,   # the AOI boundary, NOT the crop mask
  variable  = "L2-AETI-D",
  period    = wheat_period,
  folder    = wheat_download_dir,
  seasonal  = TRUE,
  mask      = TRUE
)

wheat_aeti_r <- terra::rast(wheat_aeti_seasonal_path)
wapor_plot_map(wheat_aeti_r, indicator = "Seasonal AETI (mm)", title = "Wheat Study Area — Seasonal AETI")
```

## 3.5 Isolate the Wheat Class and Harmonize the Mask

```{r}
#| label: wheat-isolate-mask
#| fig-cap: "Wheat-only mask, harmonized to the WaPOR grid"

wheat_mask_raw <- terra::classify(
  cereal_mask_raw,
  rcl = matrix(c(wheat_class_code, 1L), ncol = 2, byrow = TRUE),
  othersNA = TRUE
)

# Align to the same grid as the seasonal AETI raster we just downloaded
wheat_mask <- terra::resample(wheat_mask_raw, wheat_aeti_r, method = "near")
names(wheat_mask) <- "crop_mask"

wapor_plot_map(wheat_mask, indicator = "Wheat mask (class 1)", title = "Wheat Crop Mask")
```

## 3.6 Run the Full Seasonal Water Productivity Analysis

The indicators and equations are exactly the ones explained in §2.6 for Citrus — same
math, new crop and season.

```{r}
#| label: wheat-run-analysis

wheat_config <- list(
  period        = wheat_period,
  ref_year      = 1970,
  aeti_var      = "L2-AETI-D",      # or "L1-AETI-D" (~300 m, global) / "L3-AETI-D" (~20 m, L3 schemes only)
  ret_var       = "L1-RET-D",
  precip_var    = "L1-PCP-D",
  npp_var       = "L2-NPP-D",       # or "L1-NPP-D" (~300 m, global)
  t_var         = "L2-T-D",
  data_source   = "api",
  folder        = wheat_download_dir,
  area_weighted = TRUE,
  indicators    = all_indicators    # defined in §2.6
)

wheat_season <- wapor_run_seasonal_analysis(
  config      = wheat_config,
  crop_params = wheat_params,
  rasters     = list(crop_mask = wheat_mask),
  aoi_region  = wheat_aoi_path
)

names(wheat_season)
```

## 3.7 Explore the Outputs, Step by Step

```{r}
#| label: wheat-out-aeti-etc
#| fig-cap: "Seasonal AETI and crop water demand (ETc) for wheat"

p5 <- wapor_plot_map(wheat_season$seasonal_aeti$raster, indicator = "AETI (mm)", title = "Seasonal AETI")
wheat_etc <- wheat_season$etc_by_class[[as.character(wheat_class_code)]]$etc_seasonal
p6 <- wapor_plot_map(wheat_etc, indicator = "ETc (mm)", title = "Seasonal ETc")
patchwork::wrap_plots(p5, p6)

wheat_season$seasonal_aeti_by_class
```

```{r}
#| label: wheat-out-adequacy-bf
#| fig-cap: "Water adequacy and beneficial fraction for wheat"

p7 <- wapor_plot_map(wheat_season$adequacy_etc, indicator = "Adequacy", title = "Adequacy (AETI / ETc)")
p8 <- wapor_plot_map(wheat_season$beneficial_fraction, indicator = "T/AETI", title = "Beneficial Fraction")
patchwork::wrap_plots(p7, p8)
```

```{r}
#| label: wheat-out-green-blue
#| fig-cap: "Green (rainfall) vs. blue (irrigation) water for wheat"

p9  <- wapor_plot_map(wheat_season$green_water, indicator = "Green water (mm)", title = "Green Water")
p10 <- wapor_plot_map(wheat_season$blue_water, indicator = "Blue water (mm)", title = "Blue Water")
patchwork::wrap_plots(p9, p10)
```

A winter wheat season sown in November and harvested in May usually shows a much
higher green-water share than an irrigated summer crop, simply because it overlaps the
wet season — a useful sanity check on the numbers you get back.

```{r}
#| label: wheat-out-productivity
#| fig-cap: "Wheat yield raster (t/ha)"

wapor_plot_map(wheat_season$yield_raster, indicator = "Yield (t/ha)", title = "Estimated Wheat Yield")

data.frame(
  indicator = c("CWP (kg/m3)", "BWP (kg/m3)"),
  value     = round(c(wheat_season$cwp, wheat_season$bwp), 3)
)
```

## 3.8 Export All Outputs

```{r}
#| label: wheat-export

wapor_export_analysis_outputs(
  results      = wheat_season,
  folder       = wheat_output_dir,
  indicators   = all_indicators,
  season_label = "Wheat_Winter2023_2024"
)
```

---

# 4. Comparing the Two Crops

::: {.callout-important}
Citrus is a perennial tree crop grown over a full calendar year; wheat is an annual
cereal grown over roughly seven months in the wet season. Their $CWP$/$BWP$ values are
**not directly comparable as a measure of "which crop is more efficient"** — different
crop physiology, season length, and climate overlap all drive real differences. Treat
this chart as a way to sanity-check both runs side by side, not a ranking.
:::

```{r}
#| label: crop-comparison
#| fig-cap: "Crop and Biomass Water Productivity: Citrus vs. Wheat"

comparison_df <- data.frame(
  crop = rep(c("Citrus", "Wheat"), each = 2),
  indicator = rep(c("CWP", "BWP"), times = 2),
  value_kg_m3 = c(citrus_season$cwp, citrus_season$bwp, wheat_season$cwp, wheat_season$bwp)
)

ggplot(comparison_df, aes(x = crop, y = value_kg_m3, fill = indicator)) +
  geom_col(position = "dodge") +
  labs(
    title = "Water Productivity by Crop",
    x = NULL, y = "kg / m3", fill = NULL
  ) +
  theme_minimal()
```

---

# 5. Next Steps

- `vignette("advanced-analysis")` — custom crop tables for multiple classes at once,
  spatial anomaly/hotspot detection across years, tiled processing for large extents.
- `vignette("wheat-water-productivity")` — multi-year batch seasons and per-season
  "Smart-Linking" masks (a different wheat mask/dates for every year automatically).
- `vignette("data-catalog")` — full WaPOR v3 / AgERA5 variable table, including which
  variables exist at which spatial resolution (L1/L2/L3).
- `wapor_calc_zscore()`, `wapor_calc_spatial_hotspots()`, and `wapor_calc_anomaly_baseline()`
  — once you have more than one season of results (e.g. a multi-year wheat batch), these
  compute pixel-wise temporal anomalies and classify spatial water-stress hotspots across
  years; see `vignette("advanced-analysis")` §4 for the worked example.
