---
title: "Water Productivity Analysis with WaPOR"
subtitle: "Training manual and notebook: citrus in the Jordan Valley and wheat in Jendouba"
author: "WaPOR Training Series"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    df-print: paged
    theme: flatly
    fig-width: 8
    fig-height: 5
    fig-align: center
    embed-resources: true      # one self-contained HTML file (opens without R)
execute:
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

# How to use this manual {.unnumbered}

This notebook is both a **manual** and a **working analysis**. Read it
from top to bottom: every step explains what we are about to do and
why, runs the code, and then shows a **checkpoint** so you can confirm
the step worked before moving on.

You will learn how to use WaPOR data to answer three practical
questions for an irrigated area:

1.  **How much water did the crop consume?** (actual evapotranspiration)
2.  **Was that enough?** (adequacy, compared with crop water demand)
3.  **What did the water produce?** (biomass, yield and water
    productivity, for the annual crop; tree crop yields come from farm
    surveys)

We do this twice, on real data:

|   | Part A: Citrus | Part B: Wheat |
|------------------------|------------------------|------------------------|
| Location | North Jordan Valley, Jordan | Jendouba, Tunisia |
| Crop map | Citrus field polygons, plus 177 farms visited in the field | Cereal map (raster, 20 m) |
| Season | 1 March 2024 to 28 February 2025 | 1 November 2023 to 31 May 2024 |
| WaPOR data | Level 3 (20 m), region JVA | Level 3 (20 m), region JEN |

## What the boxes mean {.unnumbered}

::: callout-note
## Concept

Explains a term or a method. Read these the first time; come back to
them when you need a reminder.
:::

::: callout-tip
## Checkpoint

Tells you what a correct result looks like. The code also prints lines
starting with `[PASS]` or `[CHECK]`. A `[CHECK]` line does not stop the
notebook; it asks you to look more closely at that step.
:::

::: callout-warning
## Watch out

A common mistake, or something that changes the results.
:::

## Before you start {.unnumbered}

Keep the `training` folder together. The input data are already inside
it:

```
training/
├── water-productivity-training.qmd   (this notebook)
├── images/                           (the sketches used in this manual)
└── data/
    ├── citrus/
    │   ├── Citrus_NJV.geojson          citrus fields (238 polygons)
    │   ├── Citrus_farms_final.geojson  sampled farms (177 polygons, 113 farm IDs)
    │   └── survey/                     farm survey tables (CSV, no personal data)
    └── wheat/
        └── Season5_2023_2024_cereal_mask.tif   cereal map, value 1 = cereal
```

-   Open the notebook in RStudio (or Positron) from the `training`
    folder, then run the chunks one by one, or click **Render** to run
    everything.
-   The first run needs an internet connection: the WaPOR data are
    downloaded from the FAO WaPOR server (no account or key needed) and
    saved in `wapor_data/`. If that folder is already there (for example
    when the training folder was copied after a full run), the notebook
    uses the saved files and runs without internet; only the satellite
    background of the interactive maps needs a connection.
-   Keep 15 to 20 GB of free disk space for temporary files.
-   A full run takes about 35 minutes the first time, including the
    downloads; with the data already saved in `wapor_data/` it is faster.
    Downloads are saved in `wapor_data/` and results in `outputs/`.

## The workflow at a glance {.unnumbered}

Both parts follow the same main steps. Part A (citrus) adds one more:
linking the WaPOR results with a farm survey (Step 10).

![The analysis steps used in both case studies.](images/workflow.svg){width="95%"}

------------------------------------------------------------------------

# Key concepts {#sec-concepts}

This chapter explains the terms used in the rest of the manual. Each
concept is used later in the analysis, and the steps refer back here.

## What is WaPOR?

WaPOR is FAO's portal to monitor **Wa**ter **P**roductivity through
**O**pen access of **R**emotely sensed derived data. It publishes maps,
derived from satellite images, of how much water the land surface
consumes and how much biomass it produces, from 2018 to near real time.

The variables used in this manual:

| Variable | Name | Unit (as published) | What it tells you |
|------------------|------------------|------------------|------------------|
| **AETI** | Actual EvapoTranspiration and Interception | mm/day | Total water consumed (see @sec-aeti) |
| **T** | Transpiration | mm/day | Water used by the plants themselves |
| **NPP** | Net Primary Production | gC/m²/day | Carbon fixed by the plants, the basis for biomass |
| **RET** | Reference EvapoTranspiration | mm/day | How much the weather "pulls" water out of a reference grass surface |
| **PCP** | Precipitation | mm/day | Rainfall |

## Where does the water go? {#sec-aeti}

![AETI is the sum of soil evaporation (E), plant transpiration (T) and
interception (I).](images/water_balance.svg){width="90%"}

![The WaPOR variables for a citrus orchard: the water balance of a tree
(top) and how each variable looks as a map (bottom). The maps are
illustrations; the real maps for our study areas are made in Parts A
and B.](images/aeti_components.png){width="100%"}

Water that reaches a field, as rain or irrigation, leaves it again in
three ways:

-   **E, evaporation**: from the wet soil surface. It does not help the
    crop grow.
-   **T, transpiration**: through the leaves. This is the water the
    crop uses to grow, so it is called **beneficial** consumption.
-   **I, interception**: rain caught on leaves that evaporates before
    reaching the soil.

WaPOR's **AETI** is the sum $AETI = E + T + I$. The share that went
through the plants is the **beneficial fraction**, $BF = T / AETI$. A
high BF (0.7 or more for a well-managed orchard) means little water is
lost from bare, wet soil.

## Resolution: WaPOR levels {#sec-levels}

![The same 2.6 ha farm on the three WaPOR grids.](images/wapor_levels.svg){width="95%"}

WaPOR publishes AETI, T and NPP at three levels:

| Level | Pixel size | Area covered | Good for |
|------------------|------------------|------------------|------------------|
| L1 | about 300 m | Africa and Near East | Basins, regions |
| L2 | about 100 m | Africa and Near East | Large irrigation schemes |
| L3 | about 20 m | Selected irrigation schemes only | Individual farms |

Many farms in this training are only 1 to 5 ha. At L2 such a farm is
one to five pixels, most of them mixed with the neighbouring land. At
L3 the same farm is 25 to 125 pixels, so **we use L3** for both
examples. RET and precipitation are only available at L1 (about 5 km);
they vary slowly in space, so the coarse grid is acceptable.

## Dekads and seasonal totals {#sec-dekads}

![Each dekad's daily rate is multiplied by the number of days, then
the dekads are added up.](images/dekads_to_season.svg){width="90%"}

WaPOR data come as **dekads**: each month is split into three periods,
days 1 to 10, 11 to 20, and 21 to the end of the month (8 to 11 days).
Each dekadal map gives the **average daily rate** in that dekad, in
mm/day.

To get a **seasonal total** in mm, every dekad is multiplied by the
number of its days that fall inside the season, and the results are
added:

$$
AETI_{season} = \sum_{dekads} AETI_{dekad}\ (\text{mm/day}) \times \text{days in season}
$$

The same rule applies to T, RET, precipitation and NPP. Variables
that describe a state rather than a flow, such as soil moisture or
temperature, are averaged instead of added.

## Study area (AOI) and crop mask {#sec-mask}

![The AOI decides where data are downloaded; the crop mask decides
which pixels are analysed.](images/aoi_mask.svg){width="95%"}

-   The **area of interest (AOI)** is the outline of the study area. It
    tells the software which part of the WaPOR map to download.
-   The **crop mask** says which pixels inside the AOI belong to the
    crop we study. All other pixels (other crops, villages, bare land)
    are left out.
-   Before the analysis the crop mask is put on the **WaPOR grid**, so
    every WaPOR pixel is either "crop" or "not crop".

For citrus the field polygons are both the AOI and the mask. For
wheat, the AOI is the outline of the cereal map and the mask is the
map itself.

## Crop water demand: Kc and ETc {#sec-kc}

![The four-stage crop coefficient curve. Multiply it by RET to get the
crop's water demand.](images/kc_curve.svg){width="90%"}

How much water a crop **would** use if it had enough water depends on
the weather (RET) and on the crop and its growth stage. FAO Irrigation
and Drainage Paper 56 (FAO-56) describes this with a **crop
coefficient, Kc**:

$$
ET_c = K_c \times RET
$$

Kc changes over the season in four stages: **initial** (flat at
$K_{c,ini}$), **development** (rising), **mid-season** (flat at
$K_{c,mid}$) and **late season** (falling to $K_{c,end}$). The values
for each crop come from FAO-56 Table 12. $ET_c$ is the crop water
**demand**; AETI is what was actually **consumed**.

For an annual crop the four periods follow the crop from sowing to
harvest. For a tree crop they follow the tree's yearly cycle (for citrus:
flowering, fruit set, fruit development, maturation and harvest), so the
stages get the names of that cycle.

## Performance indicators {#sec-indicators}

| Indicator | Formula | How to read it |
|------------------|------------------|------------------------------------|
| **Adequacy** | $AETI / ET_c$ | Good $0.8 < A \le 1$, acceptable $0.68 < A \le 0.8$, poor $A \le 0.68$. Above 1: consumption above the crop demand. |
| **Uniformity** | $1 - CV$ of seasonal AETI of the pixels inside one field (or block) | Evenness of water use within a field. Compare with the standard for the irrigation method: drip 85%, sprinkler and centre pivot 75%, furrow 65%. |
| **Equity** | $CV$ of the mean seasonal AETI of the fields (or blocks) | Evenness of water use between fields. Good 0 to 10%, fair 10 to 25%, poor above 25%. |
| **Relative adequacy (P95)** | $AETI / P_{95}(AETI)$ | Compares each pixel with the best 5% of pixels of the same crop, no Kc needed. |
| **Beneficial fraction** | $T / AETI$ | Share of consumption that went through the crop. |
| **Effective rain** | USDA-SCS formula on monthly rain $P$: $P (125 - 0.2P) / 125$ if $P \le 250$ mm, else $125 + 0.1P$ | Part of the rain that stays in the root zone. |
| **Green water** | $\sum_{months} \min(AETI, P_{eff})$ | Consumption supplied by rain. |
| **Blue water** | $\sum_{months} \max(0, AETI - P_{eff})$ | Consumption supplied by irrigation. |
| **Crop water productivity (CWP)** | $Yield / AETI$ | kg of harvested product per m³ of water consumed. |
| **Biomass water productivity (BWP)** | $Biomass / AETI$ | kg of dry biomass per m³ of water consumed. |

Adequacy, uniformity, equity and land and water productivity follow the
WaPOR-based irrigation performance framework of Chukalla et al. (2022),
which lists the equations and the class limits used here (see
References). The same indicators can be used for any irrigated crop.

### Green and blue water are split month by month {#sec-greenblue}

![Rain in a wet month cannot supply the crop in a dry month.](images/green_blue.svg){width="95%"}

Green and blue water are calculated **for each month** and then added
up over the season. If we split the seasonal totals instead, the
surplus rain of a wet winter month would appear to "pay for" the
irrigation of a dry summer month, and the result would show far too
much green water and too little blue water. For every pixel,
$green + blue = AETI$.

### From NPP to yield and water productivity (annual crops) {#sec-yield}

This chain is used for **annual field crops** such as wheat. The factors
in it (HI, MC, fc, AOT) are calibrated for field crops. For tree crops
such as citrus, most biomass goes into wood and leaves over many years, so
yield is taken from farm surveys or records, not from this equation.

![From NPP to biomass, yield and water productivity.](images/yield_chain.svg){width="95%"}

-   **22.222** converts grams of carbon to kilograms of dry matter per
    hectare.
-   **AOT**: share of biomass above ground. **fc**: light use efficiency
    correction factor (1 for C3 crops such as wheat). **MC**: moisture
    content of the
    harvested product. **HI**: harvest index, the harvested share of
    the above-ground biomass.
-   To divide by AETI, mm are converted to m³/ha: 1 mm on 1 ha is
    10 m³.

::: callout-warning
## Yield is fresh weight, biomass is dry weight

Yield includes the water inside the harvested product (it is divided by
$1 - MC$). For wheat grain ($MC = 0.15$) this adds little, so wheat CWP
stays below BWP. Compare CWP with the fresh yields reported by farmers and
BWP with other dry-matter figures.
:::

------------------------------------------------------------------------

# Setting up

## Install the packages (first time only)

```{r}
#| label: install
#| eval: false

# The same two lines as in the note to participants
install.packages(c("remotes", "terra", "sf", "dplyr", "ggplot2", "leaflet", "knitr", "rmarkdown"))
remotes::install_github("almutaz9000/Rwapor")
```

## Load the packages

```{r}
#| label: setup

library(Rwapor)     # WaPOR download and water productivity analysis
library(terra)      # rasters
library(sf)         # vector data (polygons)
library(dplyr)      # tables
library(ggplot2)    # charts
library(leaflet)    # interactive maps

packageVersion("Rwapor")
stopifnot("Please update Rwapor (see the chunk above): version 1.0.2 or later is needed" =
            packageVersion("Rwapor") >= "1.0.2")
```

## Helper functions for the checkpoints

These small functions print the checkpoint messages and summary
statistics used after each step. You do not need to change them.

```{r}
#| label: helpers

# Print [PASS] when a condition holds, [CHECK] when it does not
check <- function(ok, what) {
  cat(if (isTRUE(ok)) "[PASS]  " else "[CHECK] ", what, "\n", sep = "")
  invisible(ok)
}

# Minimum, mean and maximum of one or more rasters
raster_stats <- function(...) {
  rs <- list(...)
  do.call(rbind, lapply(names(rs), function(n) {
    r <- rs[[n]]
    g <- terra::global(r, c("min", "mean", "max"), na.rm = TRUE)
    data.frame(layer = n, pixels = terra::global(!is.na(r), "sum")[1, 1],
               min = round(g$min, 2), mean = round(g$mean, 2), max = round(g$max, 2))
  }))
}

# Histogram of raster values, using every 20 m pixel of the crop mask
raster_hist <- function(r, xlab, ref = NULL) {
  v <- terra::values(r, na.rm = TRUE)[, 1]
  p <- ggplot(data.frame(v = v), aes(v)) +
    geom_histogram(bins = 50, fill = "#4f94cd", colour = "white") +
    labs(x = xlab, y = "Number of pixels") + theme_minimal()
  if (!is.null(ref)) p <- p + geom_vline(xintercept = ref, linetype = "dashed", colour = "red")
  p
}

# Raster map drawn with terra at full 20 m resolution (maxcell = every pixel);
# the legend title carries the unit
plot_map <- function(r, title, legend, palette = "YlGnBu", rev = TRUE) {
  terra::plot(r, main = title, col = grDevices::hcl.colors(100, palette, rev = rev),
              plg = list(title = legend), axes = FALSE, mar = c(1, 1, 2, 6),
              maxcell = terra::ncell(r))
}

# Two raster maps side by side
two_maps <- function(r1, title1, legend1, r2, title2, legend2,
                     palette1 = "YlGnBu", palette2 = palette1, rev1 = TRUE, rev2 = rev1) {
  op <- par(mfrow = c(1, 2))
  on.exit(par(op))
  plot_map(r1, title1, legend1, palette1, rev1)
  plot_map(r2, title2, legend2, palette2, rev2)
}

# Kc curve on a calendar axis, with the four Kc periods shaded and named
plot_kc_months <- function(kc_daily, start, stage_days, stage_names, title) {
  d    <- data.frame(date = as.Date(start) + seq_along(kc_daily) - 1, kc = kc_daily)
  ends <- as.Date(start) + cumsum(stage_days) - 1
  st   <- data.frame(stage = factor(stage_names, levels = stage_names),
                     from = c(as.Date(start), head(ends, -1) + 1), to = ends)
  ggplot() +
    geom_rect(data = st, aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = stage), alpha = 0.3) +
    geom_line(data = d, aes(date, kc), linewidth = 1.2, colour = "#2e7d32") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    labs(title = title, x = NULL, y = "Kc", fill = "Stage") +
    theme_minimal()
}

# Monthly means over the crop mask, for the month-by-month check
monthly_table <- function(season) {
  pick <- function(x, col, label) {
    if (is.null(x) || is.null(x$summary[[col]])) return(NULL)
    data.frame(month = as.Date(paste0(x$summary$month_key, "-01")), mm = x$summary[[col]], variable = label)
  }
  rbind(
    pick(season$monthly_aeti, "aeti_mean_mm", "AETI (consumed)"),
    pick(season$monthly_t, "t_mean_mm", "T (transpiration)"),
    pick(season$monthly_etc, "etc_mean_mm", "ETc (crop demand)"),
    pick(season$monthly_precip_peff, "pcp_mean_mm", "Rain"),
    pick(season$monthly_precip_peff, "peff_mean_mm", "Effective rain")
  )
}

plot_monthly <- function(season, title) {
  ggplot(monthly_table(season), aes(month, mm, colour = variable)) +
    geom_line(linewidth = 1) + geom_point() +
    scale_colour_manual(values = c("AETI (consumed)" = "#1565c0", "T (transpiration)" = "#2e7d32",
                                   "ETc (crop demand)" = "#e65100", "Rain" = "#90caf9",
                                   "Effective rain" = "#6a1b9a")) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    labs(title = title, x = NULL, y = "mm per month", colour = NULL) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Monthly green and blue water as stacked bars
plot_green_blue_monthly <- function(season, title) {
  g <- season$monthly_green_water$summary
  b <- season$monthly_blue_water$summary
  d <- rbind(
    data.frame(month = as.Date(paste0(g$month_key, "-01")), source = "Green (rain)", mm = g$green_mean_mm),
    data.frame(month = as.Date(paste0(b$month_key, "-01")), source = "Blue (irrigation)", mm = b$blue_mean_mm)
  )
  ggplot(d, aes(month, mm, fill = source)) +
    geom_col() +
    scale_fill_manual(values = c("Green (rain)" = "#43a047", "Blue (irrigation)" = "#1e88e5")) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    labs(title = title, x = NULL, y = "ET (mm per month)", fill = NULL) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Green and blue water with the (wrong) seasonal-total method, for comparison only
seasonal_shortcut <- function(season) list(
  green = wapor_calc_green_water(season$seasonal_aeti$raster, season$seasonal_peff),
  blue  = wapor_calc_blue_water(season$seasonal_aeti$raster, season$seasonal_peff)
)

# Download every time step (for example every dekad) of one WaPOR variable,
# cut to the study area, and save it as one GeoTIFF per date. Values are saved
# in physical units (the WaPOR scale factor is applied when reading). Files that
# are already saved are skipped, and without internet the saved files are used.
download_wapor <- function(variable, aoi, period, folder, l3_region = NULL, mask = TRUE) {
  out_dir <- file.path(folder, variable)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  urls <- tryCatch(wapor_generate_urls(variable, l3_region = l3_region, period = period),
                   error = function(e) character(0))
  if (length(urls) == 0) {
    message("No connection to the WaPOR server: using the files already saved in ", out_dir)
  }
  for (u in urls) {
    date <- wapor_date_info(u, tres = sub(".*-", "", variable))$start_date
    out  <- file.path(out_dir, sprintf("WAPOR-3.%s.%s.tif", variable, date))
    if (file.exists(out)) next
    r <- terra::rast(paste0("/vsicurl/", u))
    v <- terra::project(aoi, terra::crs(r))
    r <- terra::crop(r, v, snap = "out")
    if (mask) r <- terra::mask(r, v)
    terra::writeRaster(r, out, overwrite = TRUE, gdal = "COMPRESS=DEFLATE")
  }
  list.files(out_dir, pattern = "[.]tif$", full.names = TRUE)
}

mean_of <- function(r) if (is.null(r)) NA_real_ else round(terra::global(r, "mean", na.rm = TRUE)[1, 1], 2)
num_or_na <- function(x) if (is.null(x)) NA_real_ else round(x, 2)
```

------------------------------------------------------------------------

# Part A: Citrus in the North Jordan Valley {#sec-citrus}

## Step 1: Define the inputs

Everything specific to this study is set here. To run the same
analysis on your own data, only this chunk needs to change.

```{r}
#| label: citrus-config

# The "Setting up" chunks (packages and helper functions) must run first
stopifnot("Run the chunks in 'Setting up' first (packages and helper functions)" =
            exists("check") && exists("plot_kc_months") && exists("plot_map"))

citrus_polygon_path <- "data/citrus/Citrus_NJV.geojson"          # citrus fields: AOI and crop mask
citrus_farms_path   <- "data/citrus/Citrus_farms_final.geojson"  # farms visited in the field
citrus_period       <- c("2024-03-01", "2025-02-28")             # one full year
citrus_l3_region    <- "JVA"                                     # WaPOR L3 region: North Jordan Valley
citrus_download_dir <- "wapor_data/citrus"                       # where WaPOR downloads are saved
citrus_output_dir   <- "outputs/Citrus_2024_2025"                # where results are saved
```

::: callout-tip
## Checkpoint

Both input files must be found. The season should be 365 days long.
:::

```{r}
#| label: citrus-config-check

check(file.exists(citrus_polygon_path), paste("citrus fields found:", citrus_polygon_path))
check(file.exists(citrus_farms_path), paste("sampled farms found:", citrus_farms_path))
citrus_days <- as.integer(diff(as.Date(citrus_period))) + 1L
check(citrus_days == 365, paste("season length:", citrus_days, "days"))
```

::: callout-note
## Concept: finding your L3 region

Level 3 data exist only for selected irrigation schemes. Each scheme
has a three-letter code. This list shows all of them; **JVA** is the
North Jordan Valley and **JEN** is Jendouba.
:::

```{r}
#| label: l3-regions

# The list comes from the WaPOR server; a copy is kept for use without internet
l3_file <- "data/wapor_l3_regions.csv"
l3_regions <- tryCatch(wapor_fetch_l3_regions()[, c("code", "caption")], error = function(e) NULL)
if (is.null(l3_regions) || nrow(l3_regions) == 0) {
  l3_regions <- read.csv(l3_file)
} else {
  write.csv(l3_regions, l3_file, row.names = FALSE)
}
l3_regions
```

## Step 2: Look at the study area

We load the citrus fields and the sampled farms and show them on a
satellite image. Hover over a polygon to see its crop.

```{r}
#| label: citrus-explore-aoi
#| fig-cap: "Citrus fields (yellow) and the farms visited in the field (red)"

citrus_poly  <- sf::st_read(citrus_polygon_path, quiet = TRUE) |> sf::st_transform(4326)
citrus_farms <- sf::st_read(citrus_farms_path, quiet = TRUE)   |> sf::st_transform(4326)

leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addPolygons(data = citrus_poly, color = "#e6b800", weight = 2, fillOpacity = 0.15,
              label = ~Crop_Name) |>
  addPolygons(data = citrus_farms, color = "#d7301f", weight = 1.5, fillOpacity = 0.4,
              label = ~paste0(Name, ": ", Crop_Name)) |>
  addLegend(colors = c("#e6b800", "#d7301f"), labels = c("Citrus fields", "Sampled farm"),
            position = "bottomright")
```

```{r}
#| label: citrus-aoi-summary

# Areas are measured in the local projected system (UTM zone 36N, EPSG:32636),
# which gives areas in metres. Overlapping polygons are merged first.
area_ha <- function(x) {
  as.numeric(sf::st_area(sf::st_union(sf::st_transform(x, 32636)))) / 10000
}
citrus_area_ha <- area_ha(citrus_poly)

data.frame(
  layer      = c("Citrus fields", "Sampled farms"),
  polygons   = c(nrow(citrus_poly), nrow(citrus_farms)),
  area_ha    = round(c(citrus_area_ha, area_ha(citrus_farms)), 1)
)

# Which citrus types were sampled?
sort(table(citrus_farms$Crop_Name), decreasing = TRUE)
```

::: callout-tip
## Checkpoint

The fields should sit on the green irrigated strip east of the Jordan
river, and the sampled farms should lie inside the citrus area. Check
the list of crop names for spelling mistakes before you use them to
group results.
:::

```{r}
#| label: citrus-aoi-check

check(nrow(citrus_poly) > 0 && all(terra::is.valid(terra::vect(citrus_poly))),
      "citrus polygons loaded and valid")
inside <- sf::st_within(sf::st_transform(citrus_farms, 32636),
                        sf::st_union(sf::st_buffer(sf::st_transform(citrus_poly, 32636), 50)),
                        sparse = FALSE)[, 1]
check(mean(inside) > 0.9, sprintf("%d of %d sampled farms lie inside the citrus area", sum(inside), length(inside)))
```

## Step 3: Crop parameters for citrus

We now describe the crop's water demand with its Kc curve (@sec-kc).
Citrus is a tree crop, so we do not estimate its yield from WaPOR biomass;
only the Kc values and stage lengths are needed.

-   **Kc values** from FAO-56, [Chapter 6,
    Table 12](https://www.fao.org/4/x0490e/x0490e0b.htm#chapter%206%20%20%20etc%20%20%20single%20crop%20coefficient%20(kc))
    ("Single (time-averaged) crop coefficients, Kc"), row *Citrus, no
    ground cover, 70% canopy*: $K_{c,ini}=0.70$, $K_{c,mid}=0.65$,
    $K_{c,end}=0.70$, maximum height 4 m. If the orchards have an
    active ground cover or weeds, use the next row of the table
    instead (0.75, 0.70, 0.75).
-   **Stage lengths** for citrus in the Mediterranean, starting on
    1 March: 60, 90, 120 and 95 days (365 days in total). You give the
    first, third and fourth lengths; the second takes whatever days
    remain in your season.

::: callout-note
## Concept: the citrus year in four Kc periods

| Kc period | Days | Months (from 1 March) | Citrus growth stage |
|------------------|------------------|------------------|------------------|
| 1 (Kc ini) | 60 | March to April | Flowering and new growth |
| 2 (transition) | 90 | May to July | Fruit set and early fruit growth |
| 3 (Kc mid) | 120 | August to November | Fruit development |
| 4 (to Kc end) | 95 | December to February | Maturation and harvest |

The months are typical for Mediterranean oranges and mandarins; early and
late varieties shift by a few weeks. For citrus the Kc values change
little over the year (0.65 to 0.70), because the evergreen canopy covers
the ground all year.
:::

```{r}
#| label: citrus-crop-params

citrus_params <- wapor_create_crop_params(
  class_value  = 1L,        # the value citrus pixels get in the crop mask (Step 5)
  crop_name    = "Citrus (70% canopy, no ground cover)",
  kc_ini       = 0.70,      # FAO-56 Table 12
  kc_mid       = 0.65,
  kc_end       = 0.70,
  l_ini_days   = 60L,       # stage lengths in days
  l_mid_days   = 120L,
  l_late_days  = 95L,
  max_height_m = 4.0,       # FAO-56 Table 12
  region       = "Mediterranean"
)
citrus_params
```

Now we build the daily Kc curve for our 365-day season and plot it:

```{r}
#| label: citrus-kc-curve
#| fig-cap: "Daily Kc for citrus over the season, by citrus growth stage"

citrus_l_dev <- citrus_days - (citrus_params$l_ini_days + citrus_params$l_mid_days + citrus_params$l_late_days)

citrus_kc_daily <- wapor_build_kc(
  kc_ini = citrus_params$kc_ini, kc_mid = citrus_params$kc_mid, kc_end = citrus_params$kc_end,
  l_ini  = citrus_params$l_ini_days, l_dev = citrus_l_dev,
  l_mid  = citrus_params$l_mid_days, l_late = citrus_params$l_late_days
)

citrus_stages <- c("Flowering and new growth", "Fruit set and early growth",
                   "Fruit development", "Maturation and harvest")
plot_kc_months(citrus_kc_daily, citrus_period[1],
               c(citrus_params$l_ini_days, citrus_l_dev, citrus_params$l_mid_days, citrus_params$l_late_days),
               citrus_stages, "Citrus Kc curve, March 2024 to February 2025")
```

::: callout-tip
## Checkpoint

The curve has one value per day of the season and stays between the
Kc values you entered. For an evergreen orchard the curve is almost
flat (0.65 to 0.70); for an annual crop it rises and falls strongly
(see Part B).
:::

```{r}
#| label: citrus-kc-check

check(length(citrus_kc_daily) == citrus_days, paste("Kc curve has", length(citrus_kc_daily), "daily values"))
check(citrus_l_dev > 0, paste("development stage:", citrus_l_dev, "days"))
check(all(range(citrus_kc_daily) >= 0.65 - 1e-9 & range(citrus_kc_daily) <= 0.70 + 1e-9),
      sprintf("Kc ranges from %.2f to %.2f", min(citrus_kc_daily), max(citrus_kc_daily)))
```

## Step 4: First look at the data: seasonal AETI map

Before the full analysis we download one variable, AETI, and add its
dekads up to a seasonal total (@sec-dekads). This is a quick way to see
the data and the study area together. The map is saved, so the next time
(or without internet) the saved map is used.

```{r}
#| label: citrus-seasonal-aeti-map
#| fig-cap: "Seasonal AETI (mm) over the citrus fields, March 2024 to February 2025"

saved_map <- list.files(file.path(citrus_download_dir, "L3-AETI-D_seasonal"), "seasonal.*[.]tif$",
                        full.names = TRUE)
citrus_aeti_seasonal_path <- if (length(saved_map) > 0) saved_map[1] else wapor_map(
  region    = citrus_polygon_path,   # AOI: the citrus fields
  variable  = "L3-AETI-D",           # Level 3, dekadal AETI
  l3_region = citrus_l3_region,
  period    = citrus_period,
  folder    = citrus_download_dir,
  seasonal  = TRUE,                  # add the dekads up to one seasonal total
  mask      = TRUE                   # keep only pixels inside the polygons
)

citrus_aeti_r <- terra::rast(citrus_aeti_seasonal_path)
plot_map(citrus_aeti_r, "Citrus: seasonal AETI", "mm")
```

```{r}
#| label: citrus-seasonal-aeti-leaflet
#| fig-cap: "Interactive view of seasonal AETI. Zoom in to see individual fields."

aeti_pal <- colorNumeric("YlGnBu", terra::values(citrus_aeti_r), na.color = "transparent")

leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addRasterImage(citrus_aeti_r, colors = aeti_pal, opacity = 0.85, maxBytes = 64 * 1024^2) |>  # full 20 m
  addPolygons(data = citrus_poly, fill = FALSE, color = "black", weight = 1) |>
  addLegend(pal = aeti_pal, values = terra::values(citrus_aeti_r), title = "AETI (mm)")
```

::: callout-tip
## Checkpoint

The grid should be 20 m, and AETI should be positive everywhere. An
irrigated orchard in the Jordan Valley typically consumes about 800 to
1,300 mm per year. Very low values usually mean young, sparse or
abandoned orchards.
:::

```{r}
#| label: citrus-aeti-check
#| fig-cap: "Distribution of seasonal AETI over the citrus pixels"

raster_stats(`Seasonal AETI (mm)` = citrus_aeti_r)
check(all(round(terra::res(citrus_aeti_r)) == 20), paste("pixel size:", paste(round(terra::res(citrus_aeti_r)), collapse = " x "), "m"))
check(terra::global(citrus_aeti_r, "min", na.rm = TRUE)[1, 1] >= 0, "no negative AETI values")
raster_hist(citrus_aeti_r, "Seasonal AETI (mm)")
```

## Step 5: Put the crop mask on the WaPOR grid

The seasonal AETI map from Step 4 is on the WaPOR grid, so we use it as
a template and turn the citrus polygons into a raster with the value 1
on citrus pixels (@sec-mask).

```{r}
#| label: citrus-rasterize-mask
#| fig-cap: "Citrus crop mask (value 1) on the 20 m WaPOR grid"

# WaPOR L3 maps use a local UTM system, so the polygons are reprojected first
citrus_vect <- terra::project(terra::vect(citrus_poly), terra::crs(citrus_aeti_r))
citrus_mask <- terra::rasterize(citrus_vect, citrus_aeti_r, field = 1, background = NA)
names(citrus_mask) <- "crop_mask"

terra::plot(citrus_mask, col = "#e6b800", legend = FALSE, axes = FALSE, main = "Citrus crop mask",
            maxcell = terra::ncell(citrus_mask))
```

::: callout-tip
## Checkpoint

The area of the mask pixels should be close to the area of the
polygons (within about 10%; small differences come from pixels along
the field edges).
:::

```{r}
#| label: citrus-mask-check

mask_ha <- terra::global(!is.na(citrus_mask), "sum")[1, 1] * prod(terra::res(citrus_mask)) / 10000
data.frame(polygon_area_ha = round(citrus_area_ha, 1), mask_area_ha = round(mask_ha, 1))
check(abs(mask_ha / citrus_area_ha - 1) < 0.10,
      sprintf("mask area is %.1f%% of the polygon area", 100 * mask_ha / citrus_area_ha))
```

## Step 6: Download the data and run the seasonal analysis

This is the main step. First we download every dekad of the season for
the four variables we need and save them in `wapor_data/citrus/dekadal/`,
one file per dekad. Then the analysis reads these saved files, applies
the crop mask, and calculates the indicators (@sec-indicators).

### Download the WaPOR data (once)

| Variable | Level | Pixel size | Why we need it |
|------------------|------------------|------------------|------------------|
| AETI | L3 | 20 m | water consumption |
| T | L3 | 20 m | beneficial fraction |
| RET | L1 | coarse | crop water demand (ETc) |
| PCP | L1 | coarse | effective rain, green and blue water |

AETI and T are cut to the citrus fields. RET and precipitation are only
cut to the area around them, because one coarse L1 pixel covers many
fields. The first run downloads about 150 files (a few minutes); after
that, and without internet, the saved files are used.

```{r}
#| label: citrus-download

citrus_data_dir <- file.path(citrus_download_dir, "dekadal")
citrus_aoi_v    <- terra::vect(citrus_poly)

citrus_downloads <- list(
  AETI = download_wapor("L3-AETI-D", citrus_aoi_v, citrus_period, citrus_data_dir, citrus_l3_region),
  T    = download_wapor("L3-T-D",    citrus_aoi_v, citrus_period, citrus_data_dir, citrus_l3_region),
  RET  = download_wapor("L1-RET-D",  citrus_aoi_v, citrus_period, citrus_data_dir, mask = FALSE),
  PCP  = download_wapor("L1-PCP-D",  citrus_aoi_v, citrus_period, citrus_data_dir, mask = FALSE)
)

data.frame(variable = names(citrus_downloads),
           files    = lengths(citrus_downloads),
           size_mb  = round(sapply(citrus_downloads, function(f) sum(file.size(f))) / 1e6, 1))
check(all(lengths(citrus_downloads) == 36), "36 dekads saved for every variable")

first_aeti <- terra::rast(citrus_downloads$AETI[1])
check(terra::global(first_aeti, "max", na.rm = TRUE)[1, 1] < 15,
      "values are in mm/day (the WaPOR scale factor is applied)")
```

### Run the analysis

First we list the indicators we want. For citrus these are water
consumption, demand and adequacy; biomass and yield are not part of the
citrus analysis.

```{r}
#| label: citrus-all-indicators

citrus_indicators <- c(
  "agg_aeti", "agg_t", "agg_ret", "agg_pcp", "agg_peff",         # seasonal totals
  "etc", "adequacy_etc", "adequacy_p95", "beneficial_fraction",  # demand and performance
  "green_water", "blue_water"                                    # water source
)
```

Then we describe the analysis and run it on the saved files. It takes a
few minutes.

```{r}
#| label: citrus-run-analysis

citrus_config <- list(
  period        = citrus_period,
  aeti_var      = "L3-AETI-D",       # actual evapotranspiration, 20 m
  t_var         = "L3-T-D",          # transpiration, 20 m
  ret_var       = "L1-RET-D",        # reference ET, only at Level 1
  precip_var    = "L1-PCP-D",        # precipitation, only at Level 1
  data_source   = "local",           # read the files saved in the download step
  folder        = citrus_data_dir,
  use_crop_mask = TRUE,              # analyse only the pixels in the crop mask
  area_weighted = TRUE,              # weight averages by pixel area
  keep_intermediates = FALSE,        # do not keep every dekad in memory (saves disk space)
  indicators    = citrus_indicators
)

citrus_season <- wapor_run_seasonal_analysis(
  config      = citrus_config,
  crop_params = citrus_params,
  rasters     = list(crop_mask = citrus_mask),
  aoi_region  = citrus_polygon_path  # only download data for the citrus area
)
```

::: callout-warning
## Watch out

-   `use_crop_mask = TRUE` is required. Without it the crop mask is
    ignored and every pixel of the AOI is analysed as citrus.
-   `aoi_region` limits the analysis to your study area. With data
    streamed from the server (`data_source = "api"`), it also limits the
    downloads; without it much larger maps are downloaded, which is slow.
:::

::: callout-tip
## Checkpoint

The season has 36 dekads (12 months x 3). Every variable should have
data for every dekad (coverage 1), and the crop mask area should match
Step 5.
:::

```{r}
#| label: citrus-run-check

n_dekads <- nrow(citrus_season$dekad_table)
check(n_dekads == 36, paste("dekads in the season:", n_dekads))

citrus_season$mask_class_stats   # pixels and area of the crop mask

coverage_min <- sapply(citrus_season$coverage, function(r) {
  terra::global(terra::mask(r, citrus_season$h_mask), "min", na.rm = TRUE)[1, 1]
})
coverage_min
check(all(coverage_min >= 0.999), "every variable has data for every dekad on every crop pixel")
```

## Step 7: Month by month check

Before looking at the indicators, look at how the season developed.
This chart shows the mean over all citrus pixels, per month.

```{r}
#| label: citrus-monthly
#| fig-cap: "Monthly water consumption, crop demand and rain over the citrus fields"
#| fig-height: 5.5

plot_monthly(citrus_season, "Citrus: month by month")
```

::: callout-tip
## Checkpoint: what the chart should show

-   **Rain** falls in winter (November to March) and is close to zero
    from May to September.
-   **ETc** (demand) and **AETI** (consumption) are highest in summer,
    when the weather is hot and dry.
-   In summer AETI is far above the rain: the orchards are
    **irrigated**. If AETI dropped to near zero in summer, the mask
    would contain unirrigated land.
-   **T** is always below AETI.
:::

```{r}
#| label: citrus-monthly-check

m <- monthly_table(citrus_season)
aeti_m <- m[m$variable == "AETI (consumed)", ]
t_m    <- m[m$variable == "T (transpiration)", ]
check(all(t_m$mm <= aeti_m$mm + 1e-6), "monthly T is below AETI in every month")
summer <- format(aeti_m$month, "%m") %in% c("06", "07", "08")
check(mean(aeti_m$mm[summer]) > mean(aeti_m$mm[!summer]), "AETI is highest in summer (irrigated orchard)")
```

## Step 8: Indicators and maps

### Water consumption: AETI and transpiration

```{r}
#| label: citrus-out-aeti-t
#| fig-cap: "Seasonal AETI and transpiration (T) over the citrus fields"

two_maps(citrus_season$seasonal_aeti$raster, "Seasonal AETI", "mm",
         citrus_season$seasonal_t$raster, "Seasonal transpiration", "mm",
         palette1 = "YlGnBu", palette2 = "YlGn")

raster_stats(`AETI (mm)` = citrus_season$seasonal_aeti$raster,
             `T (mm)`    = citrus_season$seasonal_t$raster)
```

### Crop water demand and adequacy

The crop water demand $ET_c$ comes from the Kc curve (Step 3) and RET.
Adequacy compares what was consumed with that demand (@sec-indicators).

```{r}
#| label: citrus-out-etc-adequacy
#| fig-cap: "Crop water demand (ETc) and adequacy (AETI / ETc)"

citrus_etc <- citrus_season$etc_by_class[["1"]]$etc_seasonal

two_maps(citrus_etc, "Crop water demand (ETc)", "mm",
         citrus_season$adequacy_etc, "Adequacy", "AETI / ETc",
         palette1 = "Oranges", palette2 = "RdYlBu", rev1 = TRUE, rev2 = FALSE)

raster_stats(`ETc (mm)`         = citrus_etc,
             `Adequacy (ETc)`   = citrus_season$adequacy_etc,
             `Adequacy (P95)`   = citrus_season$adequacy_p95)
```

```{r}
#| label: citrus-adequacy-hist
#| fig-cap: "Distribution of adequacy. The dashed line marks 1 (consumption equals demand)."

raster_hist(citrus_season$adequacy_etc, "Adequacy (AETI / ETc)", ref = 1)

share_low <- terra::global(citrus_season$adequacy_etc < 0.8, "mean", na.rm = TRUE)[1, 1]
cat(sprintf("Share of citrus pixels with adequacy below 0.8 (possible water stress): %.0f%%\n", 100 * share_low))
```

::: callout-tip
## Checkpoint

ETc is almost the same everywhere because RET comes from a 5 km grid
and Kc is the same for all citrus. Adequacy varies with AETI. Chukalla et
al. (2022) class adequacy as good above 0.8, acceptable from 0.68 to 0.8
and poor below 0.68 (see @sec-indicators); the pixels far
below that are the ones to visit in the field.
:::

### Beneficial fraction

```{r}
#| label: citrus-out-bf
#| fig-cap: "Beneficial fraction: share of consumed water that went through the trees"

plot_map(citrus_season$beneficial_fraction, "Beneficial fraction", "T / AETI", palette = "YlGn")

bf_max <- terra::global(citrus_season$beneficial_fraction, "max", na.rm = TRUE)[1, 1]
check(bf_max <= 1 + 1e-6, sprintf("beneficial fraction never exceeds 1 (max = %.2f)", bf_max))
cat(sprintf("Mean beneficial fraction: %.2f\n", mean_of(citrus_season$beneficial_fraction)))
```

A beneficial fraction well below about 0.7 in a mature orchard means
much of the water evaporates from the soil. Mulching, drip irrigation
and less frequent wetting of the soil surface can improve it.

### Water source: green and blue water

We split consumption into rain-fed (green) and irrigation-fed (blue)
water, month by month (@sec-greenblue).

```{r}
#| label: citrus-out-green-blue
#| fig-cap: "Seasonal green water (from rain) and blue water (from irrigation)"

citrus_green <- citrus_season$green_water
citrus_blue  <- citrus_season$blue_water

two_maps(citrus_green, "Green water", "mm", citrus_blue, "Blue water", "mm",
         palette1 = "Greens", palette2 = "Blues")
```

```{r}
#| label: citrus-out-green-blue-monthly
#| fig-cap: "Monthly green and blue water (mean over the citrus fields)"

plot_green_blue_monthly(citrus_season, "Citrus: where the water came from, month by month")
```

::: callout-tip
## Checkpoint

Green plus blue must equal AETI on every pixel. Green water appears in
the rainy months, blue water dominates the dry months.
:::

```{r}
#| label: citrus-green-blue-check

gap <- terra::global(abs(citrus_green + citrus_blue - citrus_season$seasonal_aeti$raster), "max", na.rm = TRUE)[1, 1]
check(gap < 0.01, sprintf("green + blue = AETI on every pixel (largest difference %.4f mm)", gap))

# For comparison only: the wrong method that splits the seasonal totals
citrus_shortcut <- seasonal_shortcut(citrus_season)
data.frame(
  method   = c("Month by month (correct)", "Seasonal totals (wrong)"),
  green_mm = c(mean_of(citrus_green), mean_of(citrus_shortcut$green)),
  blue_mm  = c(mean_of(citrus_blue),  mean_of(citrus_shortcut$blue))
)
```

## Step 9: Results for the sampled farms

The maps describe the whole citrus area. For the farms visited in the
field we want one number per farm, to compare with the farm survey in
Step 10. For each farm we take the **area-weighted mean** of the
pixels: a pixel that only partly overlaps a farm counts only for the
overlapping part.

```{r}
#| label: citrus-farm-stats

citrus_indicator_stack <- c(
  citrus_season$seasonal_aeti$raster,
  citrus_season$seasonal_t$raster,
  citrus_etc,
  citrus_season$adequacy_etc,
  citrus_season$beneficial_fraction,
  citrus_green,
  citrus_blue
)
names(citrus_indicator_stack) <- c(
  "aeti_mm", "t_mm", "etc_mm", "adequacy", "beneficial_fraction",
  "green_mm", "blue_mm"
)

farm_stats <- terra::extract(
  citrus_indicator_stack,
  terra::project(terra::vect(citrus_farms), terra::crs(citrus_indicator_stack)),
  fun = mean, na.rm = TRUE, weights = TRUE
)

citrus_farm_results <- cbind(
  sf::st_drop_geometry(citrus_farms)[, c("Name", "Crop_Name")],
  round(farm_stats[, -1], 2)
)
citrus_farm_results
```

```{r}
#| label: citrus-farm-map
#| fig-cap: "Seasonal AETI per sampled farm. Click a farm for its values."

farms_map <- cbind(citrus_farms, citrus_farm_results[, -(1:2)])
farm_pal <- colorNumeric("RdYlBu", farms_map$aeti_mm, na.color = "grey")

leaflet(farms_map) |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addPolygons(fillColor = ~farm_pal(aeti_mm), fillOpacity = 0.8, color = "white", weight = 1,
              popup = ~sprintf("<b>%s</b> (%s)<br>AETI: %.0f mm<br>Adequacy: %.2f<br>Green: %.0f mm<br>Blue: %.0f mm",
                               Name, Crop_Name, aeti_mm, adequacy, green_mm, blue_mm)) |>
  addLegend(pal = farm_pal, values = ~aeti_mm, title = "AETI (mm)")
```

```{r}
#| label: citrus-farm-plot
#| fig-cap: "Seasonal AETI of the sampled farms, by citrus type"

ggplot(citrus_farm_results, aes(x = reorder(Crop_Name, aeti_mm, median), y = aeti_mm)) +
  geom_boxplot(fill = "#ffcc80", outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  coord_flip() +
  labs(x = NULL, y = "Seasonal AETI (mm)", title = "Water consumption of the sampled farms") +
  theme_minimal()
```

::: callout-tip
## Checkpoint

Every farm should have values. A farm with empty values (`NA`) lies
outside the citrus mask or is smaller than one pixel; find it on the
map in Step 2.
:::

```{r}
#| label: citrus-farm-check

n_na <- sum(is.na(citrus_farm_results$aeti_mm))
check(n_na == 0, sprintf("%d of %d farms have results", nrow(citrus_farm_results) - n_na, nrow(citrus_farm_results)))
```

## Step 10: Link WaPOR results with the farm survey

WaPOR tells us how much water each farm **consumed**. A farm survey tells
us how each farm is **managed** and what it **produced**. Joining the two
lets us ask practical questions: do older or denser orchards consume more
water? Do farms that apply more fertilizer or irrigation get more fruit per
m³ of water? And it gives us a **water productivity based on the real
harvest**, instead of a yield estimated from biomass.

![Joining the WaPOR farm results with the survey through the farm ID.](images/survey_join.svg){width="95%"}

::: callout-note
## Concept: the farm survey

The 113 sampled farms were interviewed in the field. The survey is stored
in four tables, all linked by the farm ID (`F001`, `F002`, ...), which is
also the `Name` of the farm polygons:

| Table | One row per | Main fields |
|------------------|------------------|------------------------------------|
| `survey_farms.csv` | farm | cultivated area, irrigation method, water source |
| `survey_crops.csv` | citrus block of a farm | citrus type, area, number of trees, yield per tree, tree age, spacing, rootstock |
| `survey_fertilizer.csv` | fertilizer product used | product, kg per year |
| `survey_irrigation.csv` | irrigation schedule | hours, interval, m³ per month |

Areas are in **dunum** (1 dunum = 1,000 m² = 0.1 ha). Personal details of
the farmers are not included in the training data.
:::

### Load the survey tables

We read the four survey tables from the training data folder and look at
the first rows of the crop table.

```{r}
#| label: citrus-survey-load

survey_dir   <- "data/citrus/survey"
survey_farms <- read.csv(file.path(survey_dir, "survey_farms.csv"))
survey_crops <- read.csv(file.path(survey_dir, "survey_crops.csv"))
survey_fert  <- read.csv(file.path(survey_dir, "survey_fertilizer.csv"))
survey_irr   <- read.csv(file.path(survey_dir, "survey_irrigation.csv"))

head(survey_crops, 8)
```

Field teams do not always type the farm ID the same way (`f002` and `F002`
are the same farm). We write every ID in capitals without spaces before
using it, and check that every surveyed farm has a polygon.

```{r}
#| label: citrus-survey-ids

norm_id <- function(x) toupper(trimws(x))
survey_ids  <- unique(norm_id(survey_farms$farm_id))
polygon_ids <- unique(norm_id(citrus_farms$Name))

data.frame(surveyed_farms = length(survey_ids), farm_polygon_ids = length(polygon_ids),
           matched = length(intersect(survey_ids, polygon_ids)))
check(all(survey_ids %in% polygon_ids), "every surveyed farm has a polygon")
```

::: callout-tip
## Checkpoint

All 113 surveyed farms match a farm ID in the polygons. A farm can have
several polygons (177 polygons for 113 farms), so the WaPOR values are
summarised per farm in the next steps.
:::

### Clean the crop table and summarise it per farm

The crop table has one row per citrus block. Before summarising we:

-   keep citrus only (a few farms also grow banana, guava, mango or palms),
-   correct an obvious spelling mistake (`Pomelit` is `Pomelo`),
-   turn the yield into a number: a dash (`-`) means young trees that do
    not bear fruit yet, which becomes "no value" (`NA`).

Then, for each farm:

| Farm value | How it is calculated |
|------------------|------------------------------------|
| Citrus area (ha) | sum of block areas / 10 |
| Tree density (trees/ha) | number of trees / citrus area |
| Tree age (years) | average age of the blocks, weighted by number of trees |
| Main citrus type | type with the largest area |
| Production (t) | sum over bearing blocks of trees x yield per tree / 1000 |
| Yield (t/ha) | production / area of the bearing blocks |

```{r}
#| label: citrus-survey-crops

citrus_types <- c("Lemon", "Navel Orange", "Shamouti Orange", "Valencia Orange", "Blood Orange",
                  "French Orange", "Local Orange", "Clementine", "Mandarin", "Grapefruit", "Pomelo")

crops <- survey_crops |>
  mutate(farm_id = norm_id(farm_id),
         crop = trimws(crop),
         crop = ifelse(crop == "Pomelit", "Pomelo", crop),
         yield_kg_per_tree = suppressWarnings(as.numeric(yield_kg_per_tree))) |>
  filter(crop %in% citrus_types)

cat("Citrus blocks kept:", nrow(crops), "of", nrow(survey_crops), "survey rows\n")
sort(table(crops$crop), decreasing = TRUE)

farm_survey <- crops |>
  group_by(farm_id) |>
  summarise(
    citrus_area_ha  = sum(area_dunum, na.rm = TRUE) / 10,
    tree_age_years  = weighted.mean(tree_age_years, trees, na.rm = TRUE),
    n_trees         = sum(trees, na.rm = TRUE),
    tree_density_ha = n_trees / citrus_area_ha,
    main_type       = crop[which.max(area_dunum)],
    bearing_area_ha = sum(area_dunum[!is.na(yield_kg_per_tree)], na.rm = TRUE) / 10,
    production_t    = if (any(!is.na(yield_kg_per_tree))) sum(trees * yield_kg_per_tree, na.rm = TRUE) / 1000 else NA_real_,
    .groups = "drop"
  ) |>
  mutate(yield_t_ha = ifelse(bearing_area_ha > 0, production_t / bearing_area_ha, NA))

head(farm_survey)
```

Fertilizer, irrigation and the irrigation method are added per farm. The
irrigation method was typed in many ways ("Open pipes Drip", "virojet",
"-"), so we group the answers: open pipes, drip, Virojet (a
micro-sprinkler), surface, and "Mixed" for farms that named more than one
method. The fertilizer table lists
products (ammonium sulfate, NPK mixtures, ...), so the total is **kg of
product**, not kg of nutrient. The irrigation table gives m³ per month;
times 12 gives m³ per year, and dividing by the area gives a depth in mm
that can be compared with AETI (1 mm on 1 ha = 10 m³).

```{r}
#| label: citrus-survey-inputs

fert_farm <- survey_fert |>
  mutate(farm_id = norm_id(farm_id),
         kg = suppressWarnings(as.numeric(quantity_kg_per_year))) |>
  group_by(farm_id) |>
  summarise(fertilizer_kg = sum(kg, na.rm = TRUE), .groups = "drop")

irr_farm <- survey_irr |>
  mutate(farm_id = norm_id(farm_id)) |>
  group_by(farm_id) |>
  summarise(applied_m3_year = 12 * sum(irrigation_m3_per_month, na.rm = TRUE), .groups = "drop")

farm_survey <- farm_survey |>
  left_join(survey_farms |> mutate(farm_id = norm_id(farm_id)) |>
              select(farm_id, cultivated_area_dunum, irrigation_method, water_source),
            by = "farm_id") |>
  left_join(fert_farm, by = "farm_id") |>
  left_join(irr_farm, by = "farm_id") |>
  mutate(method = tolower(trimws(irrigation_method)),
         irrigation_method = case_when(
           method %in% c("", "-")  ~ NA_character_,
           method == "open pipes"  ~ "Open pipes",
           method == "drip"        ~ "Drip",
           method == "virojet"     ~ "Virojet",
           method == "surface"     ~ "Surface",
           TRUE                    ~ "Mixed"),
         cultivated_ha         = cultivated_area_dunum / 10,
         fertilizer_kg_ha      = fertilizer_kg / cultivated_ha,
         applied_irrigation_mm = applied_m3_year / cultivated_ha / 10)

farm_survey <- select(farm_survey, -method)
table(farm_survey$irrigation_method, useNA = "ifany")
summary(farm_survey[, c("tree_density_ha", "tree_age_years", "yield_t_ha",
                        "fertilizer_kg_ha", "applied_irrigation_mm")])
```

::: callout-tip
## Checkpoint

Look at the minimum and maximum of each column. Typical values (to
confirm with the survey team): 200 to 700 trees/ha, 5 to 50 years, 15 to
60 t/ha of fruit, 500 to 1,500 mm of applied irrigation per year. Values
far outside these ranges usually come from typing errors in the survey
(wrong unit, a missing zero) and should be checked before drawing
conclusions.
:::

The next cell counts the farms and lists any farm whose yield falls
outside a wide plausible range, so it can be checked with the survey team.

```{r}
#| label: citrus-survey-check

check(nrow(farm_survey) == length(survey_ids),
      sprintf("%d farms summarised from the survey", nrow(farm_survey)))
odd_yield <- farm_survey$farm_id[!is.na(farm_survey$yield_t_ha) &
                                   (farm_survey$yield_t_ha < 5 | farm_survey$yield_t_ha > 80)]
check(length(odd_yield) == 0,
      paste("farm yields between 5 and 80 t/ha",
            if (length(odd_yield)) paste0("(check: ", paste(odd_yield, collapse = ", "), ")")))
cat("Farms with a known yield:", sum(!is.na(farm_survey$yield_t_ha)), "\n")
```

### WaPOR values per farm

Some farms have several polygons. We first merge the polygons of each farm
into one outline, then take the area-weighted mean of every WaPOR
indicator inside it (the same method as Step 9).

```{r}
#| label: citrus-survey-wapor

farms_by_id <- terra::project(terra::vect(citrus_farms), terra::crs(citrus_indicator_stack)) |>
  terra::aggregate(by = "Name")
wapor_farm <- terra::extract(citrus_indicator_stack, farms_by_id, fun = mean, na.rm = TRUE, weights = TRUE)
wapor_farm$farm_id <- norm_id(farms_by_id$Name)
wapor_farm <- wapor_farm[, c("farm_id", names(citrus_indicator_stack))]

head(wapor_farm)
check(nrow(wapor_farm) == length(polygon_ids), sprintf("%d farms with WaPOR values", nrow(wapor_farm)))
```

### Join the two tables

The farm ID links the WaPOR values and the survey answers into one farm
table.

```{r}
#| label: citrus-survey-join

farm_table <- inner_join(wapor_farm, farm_survey, by = "farm_id")
check(nrow(farm_table) == length(survey_ids),
      sprintf("%d farms have both WaPOR values and survey answers", nrow(farm_table)))
```

### Water productivity from the surveyed yield

Now we can calculate water productivity from the fruit that was really
harvested:

$$
CWP_{survey}\ (\text{kg/m}^3) = \frac{Yield_{survey}\ (\text{kg/ha})}{AETI\ (\text{mm}) \times 10}
$$

Two more farm indicators help to compare orchards of different density:
the **water consumed per tree** (m³/tree) and the **fruit per tree** (kg).

```{r}
#| label: citrus-survey-cwp

farm_table <- farm_table |>
  mutate(cwp_kg_m3        = (yield_t_ha * 1000) / (aeti_mm * 10),
         aeti_m3_per_tree = (aeti_mm * 10) / tree_density_ha,
         fruit_kg_per_tree = production_t * 1000 / n_trees)

farm_table |>
  select(farm_id, main_type, aeti_mm, yield_t_ha, cwp_kg_m3, aeti_m3_per_tree, fruit_kg_per_tree) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  head(10)

summary(farm_table$cwp_kg_m3)
```

The box plot compares water productivity between the main citrus types.

```{r}
#| label: citrus-survey-cwp-type
#| fig-cap: "Water productivity from the surveyed yield, by main citrus type"

ggplot(filter(farm_table, !is.na(cwp_kg_m3)),
       aes(x = reorder(main_type, cwp_kg_m3, median), y = cwp_kg_m3)) +
  geom_boxplot(fill = "#a5d6a7", outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
  coord_flip() +
  labs(x = NULL, y = "CWP (kg of fruit per m³ of water consumed)",
       title = "Water productivity of the sampled farms") +
  theme_minimal()
```

The map shows where the more and less productive farms are.

```{r}
#| label: citrus-survey-cwp-map
#| fig-cap: "Water productivity per farm (surveyed yield / WaPOR AETI). Click a farm for its values."

farm_map <- sf::st_transform(sf::st_as_sf(farms_by_id), 4326)
farm_map$farm_id <- norm_id(farm_map$Name)
farm_map <- merge(farm_map, farm_table, by = "farm_id")
cwp_pal <- colorNumeric("RdYlGn", farm_map$cwp_kg_m3, na.color = "grey")

leaflet(farm_map) |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addPolygons(fillColor = ~cwp_pal(cwp_kg_m3), fillOpacity = 0.85, color = "white", weight = 1,
              popup = ~sprintf("<b>%s</b> (%s)<br>AETI: %.0f mm<br>Yield: %.1f t/ha<br>CWP: %.2f kg/m³<br>Trees: %.0f /ha, %.0f years",
                               farm_id, main_type, aeti_mm, yield_t_ha, cwp_kg_m3, tree_density_ha, tree_age_years)) |>
  addLegend(pal = cwp_pal, values = ~cwp_kg_m3, title = "CWP (kg/m³)")
```

::: callout-tip
## Checkpoint

Only farms that reported a yield per tree have a CWP (in this survey
about a third of the farms); the others are grey on the map. Median CWP
around 2 to 3 kg of fruit per m³ is typical for irrigated citrus. Very high or very low CWP values usually trace back to the
yield reported in the survey; check those farms with the survey team
before using them.
:::

### How management relates to water use and productivity

We plot each farm's water consumption (AETI) against the management data.
The black line is a simple linear trend with its uncertainty band.

```{r}
#| label: citrus-survey-drivers
#| fig-cap: "Water consumption (AETI) against management data, one point per farm"
#| fig-height: 6

drivers <- c(tree_density_ha       = "Tree density (trees/ha)",
             tree_age_years        = "Tree age (years)",
             fertilizer_kg_ha      = "Fertilizer (kg product/ha/year)",
             applied_irrigation_mm = "Applied irrigation (mm/year)")

drivers_long <- do.call(rbind, lapply(names(drivers), function(v) {
  data.frame(driver = drivers[[v]], x = farm_table[[v]],
             aeti_mm = farm_table$aeti_mm, cwp_kg_m3 = farm_table$cwp_kg_m3)
}))

ggplot(filter(drivers_long, !is.na(x)), aes(x, aeti_mm)) +
  geom_point(alpha = 0.6, colour = "#1565c0") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "black", linewidth = 0.6) +
  facet_wrap(~ driver, scales = "free_x") +
  labs(x = NULL, y = "Seasonal AETI (mm)") +
  theme_minimal()
```

The same comparison for water productivity:

```{r}
#| label: citrus-survey-drivers-cwp
#| fig-cap: "Water productivity (from surveyed yield) against management data"
#| fig-height: 6

ggplot(filter(drivers_long, !is.na(x), !is.na(cwp_kg_m3)), aes(x, cwp_kg_m3)) +
  geom_point(alpha = 0.6, colour = "#2e7d32") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "black", linewidth = 0.6) +
  facet_wrap(~ driver, scales = "free_x") +
  labs(x = NULL, y = "CWP (kg/m³)") +
  theme_minimal()
```

::: callout-note
## Concept: reading a relation with a correlation coefficient

The table gives the **Spearman correlation** ($\rho$) between each
management variable and each result, with the number of farms ($n$) and
the p-value. $\rho$ goes from -1 to 1: close to 0 means no consistent
relation; positive means both rise together. Spearman uses ranks, so a
few extreme farms do not dominate it. A p-value below 0.05 suggests the
relation is unlikely to be chance.
:::

```{r}
#| label: citrus-survey-correlation

results_vars <- c(aeti_mm = "AETI", adequacy = "Adequacy", yield_t_ha = "Yield", cwp_kg_m3 = "CWP")

cor_table <- do.call(rbind, lapply(names(drivers), function(v) {
  do.call(rbind, lapply(names(results_vars), function(y) {
    ok <- complete.cases(farm_table[[v]], farm_table[[y]])
    ct <- suppressWarnings(cor.test(farm_table[[v]][ok], farm_table[[y]][ok],
                                    method = "spearman", exact = FALSE))
    data.frame(management = drivers[[v]], result = results_vars[[y]], n = sum(ok),
               rho = round(unname(ct$estimate), 2), p_value = signif(ct$p.value, 2))
  }))
}))
cor_table
```

Categories such as the citrus type or the irrigation method are compared
with box plots:

```{r}
#| label: citrus-survey-groups
#| fig-cap: "Water consumption by main citrus type and by irrigation method"
#| fig-height: 5.5

group_long <- rbind(
  data.frame(grouping = "Main citrus type", group = farm_table$main_type, aeti_mm = farm_table$aeti_mm),
  data.frame(grouping = "Irrigation method", group = farm_table$irrigation_method, aeti_mm = farm_table$aeti_mm)
)
ggplot(filter(group_long, !is.na(group)), aes(x = group, y = aeti_mm)) +
  geom_boxplot(fill = "#90caf9", outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  coord_flip() +
  facet_wrap(~ grouping, scales = "free_y") +
  labs(x = NULL, y = "Seasonal AETI (mm)") +
  theme_minimal()
```

### Applied irrigation compared with blue water consumption

The survey tells us how much irrigation water was **applied**; WaPOR tells
us how much irrigation water was **consumed** (blue ET, Step 8). Their
ratio shows how much of the applied water the trees used:

$$
\text{Share consumed} = \frac{Blue\ ET\ (\text{mm})}{Applied\ irrigation\ (\text{mm})}
$$

Values well below 1 mean much of the applied water was not consumed (it
drained, ran off or was lost from the network). Values above 1 are not
possible from irrigation alone: they point to other water sources
(groundwater, a shallow water table) or to under-reported irrigation in
the survey.

```{r}
#| label: citrus-survey-irrigation
#| fig-cap: "Blue water consumed (WaPOR) against irrigation applied (survey). The dashed line is 1:1."

farm_table$share_consumed <- farm_table$blue_mm / farm_table$applied_irrigation_mm

ggplot(filter(farm_table, !is.na(share_consumed)), aes(applied_irrigation_mm, blue_mm)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(aes(colour = irrigation_method), size = 2, alpha = 0.8) +
  labs(x = "Applied irrigation, survey (mm/year)", y = "Blue ET, WaPOR (mm)", colour = "Irrigation method") +
  theme_minimal()

summary(farm_table$share_consumed)
cat(sprintf("Farms consuming more blue water than the survey says they apply: %.0f%%\n",
            100 * mean(farm_table$share_consumed > 1, na.rm = TRUE)))
```

::: callout-tip
## Checkpoint: what this comparison tells us

In this survey most farms lie **above** the 1:1 line: WaPOR measures more
blue water consumption than the irrigation volumes the farmers reported
(median share about 1.5). Consumption cannot exceed the water supplied,
so the reported volumes are most likely too low (for example, volumes
given for part of the year, or shared pumps and ponds). This is a common
result: survey water volumes are rough, and WaPOR is a useful
cross-check. Treat the applied irrigation from the survey with caution in
the relations above.
:::

::: callout-warning
## Watch out: what the survey can and cannot tell us

-   Survey answers are estimates given by farmers from memory; yields and
    irrigation volumes can be off by a large margin. Check extreme values
    before using them.
-   Most farms grow several citrus types. The farm results mix them; the
    main type is only a label.
-   A correlation is not a cause. Older orchards, for example, are often
    also larger and on different soils.
-   With about 100 farms, weak relations ($|\rho| < 0.3$) are hard to
    separate from chance.
:::

Finally we save the joined farm table, so it can be used in Excel, QGIS or
statistics software:

```{r}
#| label: citrus-survey-save

dir.create(citrus_output_dir, recursive = TRUE, showWarnings = FALSE)
farm_table_path <- file.path(citrus_output_dir, "Citrus_2024_2025_farms_wapor_survey.csv")
write.csv(farm_table, farm_table_path, row.names = FALSE)
check(file.exists(farm_table_path), paste("farm table saved:", farm_table_path))
```

## Step 11: Save the results

All maps (GeoTIFF) and tables (CSV) are written to the output folder,
so you can open them in QGIS or Excel.

```{r}
#| label: citrus-export

wapor_export_analysis_outputs(
  results    = list(Citrus_2024_2025 = citrus_season),  # the name becomes the sub-folder
  folder     = dirname(citrus_output_dir),
  indicators = citrus_indicators,
  include_dekadal = FALSE       # the dekadal inputs are already saved in wapor_data/
)
write.csv(citrus_farm_results, file.path(citrus_output_dir, "Citrus_2024_2025_sampled_farms.csv"), row.names = FALSE)

citrus_files <- list.files(citrus_output_dir, recursive = TRUE)
check(length(citrus_files) > 0, paste(length(citrus_files), "files written to", citrus_output_dir))
table(folder = dirname(citrus_files))
```

------------------------------------------------------------------------

# Part B: Wheat in Jendouba {#sec-wheat}

Part B repeats the same steps for a winter cereal. The differences are
explained where they appear: the crop map is a **raster** instead of
polygons, the AOI is taken from that raster, and the crop parameters
come from a built-in FAO profile.

## Step 1: Define the inputs

```{r}
#| label: wheat-config

# The "Setting up" chunks (packages and helper functions) must run first
stopifnot("Run the chunks in 'Setting up' first (packages and helper functions)" =
            exists("check") && exists("plot_kc_months") && exists("plot_map"))

cereal_mask_path   <- "data/wheat/Season5_2023_2024_cereal_mask.tif"  # cereal map, 2023/24 season
wheat_class_code   <- 1L                              # value of wheat pixels in the map
wheat_period       <- c("2023-11-01", "2024-05-31")   # sowing to harvest
wheat_l3_region    <- "JEN"                           # WaPOR L3 region: Jendouba
wheat_download_dir <- "wapor_data/wheat"
wheat_output_dir   <- "outputs/Wheat_Winter2023_2024"
wheat_aoi_path     <- file.path(wheat_download_dir, "wheat_aoi.geojson")  # created in Step 2

dir.create(wheat_download_dir, recursive = TRUE, showWarnings = FALSE)
check(file.exists(cereal_mask_path), paste("cereal map found:", cereal_mask_path))
wheat_days <- as.integer(diff(as.Date(wheat_period))) + 1L
cat("Season length:", wheat_days, "days\n")
```

## Step 2: Look at the cereal map and create the AOI

```{r}
#| label: wheat-explore-inputs

cereal_mask_raw <- terra::rast(cereal_mask_path)
cereal_mask_raw

# Which values does the map contain, and how many pixels of each?
terra::freq(cereal_mask_raw)
```

::: callout-tip
## Checkpoint

Check that `wheat_class_code` in Step 1 matches the value of wheat in
the legend of your crop map. This map has a single value, 1 (cereal),
which we treat as wheat. It is 20 m in UTM zone 32N.
:::

There is no separate outline file for this study, so we use the
**outline of the cereal map** as the AOI and save it as a file:

```{r}
#| label: wheat-derive-aoi

wheat_aoi <- terra::as.polygons(terra::ext(cereal_mask_raw), crs = terra::crs(cereal_mask_raw)) |>
  sf::st_as_sf() |>
  sf::st_transform(4326)
sf::st_write(wheat_aoi, wheat_aoi_path, delete_dsn = TRUE, quiet = TRUE)

cereal_ha <- terra::freq(cereal_mask_raw)$count[1] * prod(terra::res(cereal_mask_raw)) / 10000
data.frame(
  aoi_area_ha    = round(as.numeric(sf::st_area(sf::st_transform(wheat_aoi, 32632))) / 10000),
  cereal_area_ha = round(cereal_ha)
)
check(file.exists(wheat_aoi_path), paste("AOI saved:", wheat_aoi_path))
```

```{r}
#| label: wheat-leaflet-inputs
#| fig-cap: "The AOI (white) and the cereal map (orange)"

# The map shows the cereal mask at its full 20 m resolution

leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addRasterImage(cereal_mask_raw, colors = "#f4a300", opacity = 0.7, method = "ngb",
                 maxBytes = 64 * 1024^2) |>
  addPolygons(data = wheat_aoi, fill = FALSE, color = "white", weight = 2) |>
  addLegend(colors = c("#f4a300", "white"), labels = c("Cereal", "AOI"), position = "bottomright")
```

## Step 3: Crop parameters for wheat

For wheat we can start from the built-in FAO-56 profile and change
only what we know locally. To see all built-in crops, run
`wapor_list_crops()`. Besides the Kc values, wheat needs the four yield
factors explained at the end of this step (HI, AOT, fc, MC).

```{r}
#| label: wheat-crop-params

wheat_params <- wapor_custom_crop(
  base_crop   = "Winter Wheat",     # FAO-56 profile
  class_value = wheat_class_code,
  crop_name   = "Irrigated Winter Wheat",
  kc_ini      = 0.35,
  kc_mid      = 1.15,
  kc_end      = 0.30,
  HI          = 0.43,               # harvest index: share of above-ground biomass harvested as grain
  AOT         = 0.75,               # above-ground over total biomass ratio
  fc          = 1.0,                # light use efficiency correction factor (1 for C3 crops such as wheat)
  MC          = 0.15                # moisture content of the harvested grain
)
wheat_params
```

```{r}
#| label: wheat-kc-curve
#| fig-cap: "Daily Kc for winter wheat over the season"

wheat_l_dev <- wheat_days - (wheat_params$l_ini_days + wheat_params$l_mid_days + wheat_params$l_late_days)

wheat_kc_daily <- wapor_build_kc(
  kc_ini = wheat_params$kc_ini, kc_mid = wheat_params$kc_mid, kc_end = wheat_params$kc_end,
  l_ini  = wheat_params$l_ini_days, l_dev = wheat_l_dev,
  l_mid  = wheat_params$l_mid_days, l_late = wheat_params$l_late_days
)

plot_kc_months(wheat_kc_daily, wheat_period[1],
               c(wheat_params$l_ini_days, wheat_l_dev, wheat_params$l_mid_days, wheat_params$l_late_days),
               c("Initial (sowing, emergence)", "Development (tillering)",
                 "Mid-season (heading, grain filling)", "Late season (ripening)"),
               "Winter wheat Kc curve, November 2023 to May 2024")

check(length(wheat_kc_daily) == wheat_days, paste("Kc curve has", length(wheat_kc_daily), "daily values"))
check(wheat_l_dev > 0, paste("development stage:", wheat_l_dev, "days"))
```

::: callout-note
## Concept: why the wheat curve looks so different

Wheat is an annual crop: the field is almost bare after sowing (low
Kc), the canopy closes in late winter (high Kc) and the crop dries out
before harvest (low Kc again). The evergreen citrus canopy in Part A
keeps a nearly constant Kc all year.
:::

### The yield equation and the wheat factors

WaPOR does not measure yield directly. It measures **NPP**, the carbon
the crop fixed. The yield equation turns the seasonal NPP into total
biomass and then into harvestable grain:

![Yield estimation equation for wheat, and the meaning of each
factor.](images/wheat_yield_equation.png){width="100%"}

$$
TBP = \frac{NPP \times 22.222}{1000}
$$

$$
Y = HI \times AOT \times fc \times \frac{TBP}{1 - MC}
$$

| Term | Meaning | Wheat value |
|------------------|------------------------------------|------------------|
| NPP | Net primary production over the season, from WaPOR (gC/m²) | from WaPOR |
| TBP | Total biomass production (t DM/ha); 22.222 converts gC/m² to kg DM/ha, / 1000 gives t/ha | calculated |
| fc | Light use efficiency correction factor; 1 for C3 crops such as wheat (no correction) | 1.0 |
| AOT | Above-ground over total biomass ratio: 75% of the biomass is above ground (stems, leaves, ears) | 0.75 |
| HI | Harvest index: share of the above-ground biomass harvested as grain | 0.43 |
| MC | Moisture content of the harvested grain; dividing by (1 - MC) gives yield as harvested (fresh) weight | 0.15 |
| Y | Harvestable yield (t/ha) | calculated |

The next cell prints the factors stored in `wheat_params` and confirms
they are the values in the figure, so the yields in Step 8 use exactly
these factors.

```{r}
#| label: wheat-yield-factors

yield_factors <- data.frame(
  factor  = c("HI", "AOT", "fc", "MC"),
  meaning = c("harvest index", "above-ground over total biomass",
              "light use efficiency correction (C3)", "moisture content of grain"),
  value   = c(wheat_params$HI, wheat_params$AOT, wheat_params$fc, wheat_params$MC)
)
yield_factors

check(isTRUE(all.equal(yield_factors$value, c(0.43, 0.75, 1.0, 0.15))),
      "wheat yield factors match the yield equation figure (HI 0.43, AOT 0.75, fc 1.0, MC 0.15)")

# One worked example: a pixel with 500 gC/m2 of seasonal NPP
npp_example <- 500
tbp_example <- npp_example * 22.222 / 1000
y_example   <- wheat_params$HI * wheat_params$AOT * wheat_params$fc * tbp_example / (1 - wheat_params$MC)
cat(sprintf("NPP %d gC/m2 -> TBP %.1f t DM/ha -> harvestable yield %.2f t/ha\n",
            npp_example, tbp_example, y_example))
```

## Step 4: First look: seasonal AETI map

```{r}
#| label: wheat-seasonal-aeti-map
#| fig-cap: "Seasonal AETI (mm) over the Jendouba AOI, November 2023 to May 2024"

# Use the saved map if it exists (next runs, or no internet)
saved_map <- list.files(file.path(wheat_download_dir, "L3-AETI-D_seasonal"), "seasonal.*[.]tif$",
                        full.names = TRUE)
wheat_aeti_seasonal_path <- if (length(saved_map) > 0) saved_map[1] else wapor_map(
  region    = wheat_aoi_path,       # the AOI outline, not the crop map
  variable  = "L3-AETI-D",
  l3_region = wheat_l3_region,
  period    = wheat_period,
  folder    = wheat_download_dir,
  seasonal  = TRUE,
  mask      = TRUE
)

wheat_aeti_r <- terra::rast(wheat_aeti_seasonal_path)
plot_map(wheat_aeti_r, "Jendouba: seasonal AETI", "mm")

raster_stats(`Seasonal AETI (mm)` = wheat_aeti_r)
```

::: callout-tip
## Checkpoint

This map covers the whole AOI, not only wheat: forests and permanent
crops appear with higher AETI, bare and urban land with lower values.
Winter cereals in northern Tunisia typically consume 250 to 450 mm over
the season.
:::

## Step 5: Put the crop mask on the WaPOR grid

::: callout-note
## Concept: matching a crop map to the WaPOR grid

A crop map made by another project has its own grid, and it has to be
put on the WaPOR grid before the analysis. We do it in a way that works
for any pair of grids:

1.  set wheat = 1 and everything else = 0,
2.  average these values inside each WaPOR pixel, which gives the
    **wheat fraction** of that pixel (0 to 1),
3.  keep the WaPOR pixels that are **at least 50% wheat**
    (`wheat_min_fraction`).

In Jendouba the cereal map and the WaPOR data share exactly the same
20 m grid, so every fraction is 0 or 1 and every cereal pixel is kept.
:::

```{r}
#| label: wheat-isolate-mask
#| fig-cap: "Wheat crop mask on the WaPOR grid"

wheat_min_fraction <- 0.5

# 1 = wheat, 0 = anything else
wheat_binary <- terra::ifel(!is.na(cereal_mask_raw) & cereal_mask_raw == wheat_class_code, 1, 0)

# Wheat fraction of each WaPOR pixel
wheat_fraction <- terra::project(wheat_binary, wheat_aeti_r, method = "average")

wheat_mask <- terra::ifel(wheat_fraction >= wheat_min_fraction, wheat_class_code, NA)
names(wheat_mask) <- "crop_mask"

terra::plot(wheat_mask, col = "#f4a300", legend = FALSE, axes = FALSE, main = "Wheat crop mask",
            maxcell = terra::ncell(wheat_mask))
```

```{r}
#| label: wheat-mask-check

cereal_pixels <- terra::freq(cereal_mask_raw)$count[1]
kept_pixels   <- terra::global(!is.na(wheat_mask), "sum")[1, 1]
data.frame(cereal_pixels_in_map = cereal_pixels, wapor_pixels_kept = kept_pixels)
check(abs(kept_pixels / cereal_pixels - 1) < 0.01, "the mask keeps all cereal pixels (same grid)")
```

## Step 6: Download the data and run the seasonal analysis

The method is the same as for citrus (Part A, Step 6), with one addition:
for wheat we also estimate biomass, yield and water productivity from NPP
(@sec-yield), so NPP is downloaded as well.

### Download the WaPOR data (once)

The season has 21 dekads. AETI, T and NPP (20 m) are cut to the AOI,
RET and precipitation (L1) to the area around it. The first run downloads
about 105 files; after that, and without internet, the saved files in
`wapor_data/wheat/dekadal/` are used.

```{r}
#| label: wheat-download

wheat_data_dir <- file.path(wheat_download_dir, "dekadal")
wheat_aoi_v    <- terra::vect(wheat_aoi)

wheat_downloads <- list(
  AETI = download_wapor("L3-AETI-D", wheat_aoi_v, wheat_period, wheat_data_dir, wheat_l3_region),
  T    = download_wapor("L3-T-D",    wheat_aoi_v, wheat_period, wheat_data_dir, wheat_l3_region),
  NPP  = download_wapor("L3-NPP-D",  wheat_aoi_v, wheat_period, wheat_data_dir, wheat_l3_region),
  RET  = download_wapor("L1-RET-D",  wheat_aoi_v, wheat_period, wheat_data_dir, mask = FALSE),
  PCP  = download_wapor("L1-PCP-D",  wheat_aoi_v, wheat_period, wheat_data_dir, mask = FALSE)
)

data.frame(variable = names(wheat_downloads),
           files    = lengths(wheat_downloads),
           size_mb  = round(sapply(wheat_downloads, function(f) sum(file.size(f))) / 1e6, 1))
check(all(lengths(wheat_downloads) == 21), "21 dekads saved for every variable")
```

### Run the analysis

```{r}
#| label: wheat-indicators

wheat_indicators <- c(
  citrus_indicators,                                                    # consumption, demand, water source
  "agg_npp", "agg_biomass_kg", "agg_biomass_t", "yield_npp", "cwp_bwp"  # production
)
```

Now we describe the analysis and run it.

```{r}
#| label: wheat-run-analysis

wheat_config <- list(
  period        = wheat_period,
  aeti_var      = "L3-AETI-D",
  t_var         = "L3-T-D",
  npp_var       = "L3-NPP-D",
  ret_var       = "L1-RET-D",
  precip_var    = "L1-PCP-D",
  data_source   = "local",          # read the files saved in the download step
  folder        = wheat_data_dir,
  use_crop_mask = TRUE,             # required: otherwise every AOI pixel counts as wheat
  area_weighted = TRUE,
  keep_intermediates = FALSE,       # saves disk space
  indicators    = wheat_indicators
)

wheat_season <- wapor_run_seasonal_analysis(
  config      = wheat_config,
  crop_params = wheat_params,
  rasters     = list(crop_mask = wheat_mask),
  aoi_region  = wheat_aoi_path
)
```

```{r}
#| label: wheat-run-check

n_dekads <- nrow(wheat_season$dekad_table)
check(n_dekads == 21, paste("dekads in the season:", n_dekads, "(7 months x 3)"))

wheat_season$mask_class_stats

coverage_min <- sapply(wheat_season$coverage, function(r) {
  terra::global(terra::mask(r, wheat_season$h_mask), "min", na.rm = TRUE)[1, 1]
})
coverage_min
check(all(coverage_min >= 0.999), "every variable has data for every dekad on every wheat pixel")
```

## Step 7: Month by month check

```{r}
#| label: wheat-monthly
#| fig-cap: "Monthly water consumption, crop demand and rain over the wheat fields"
#| fig-height: 5.5

plot_monthly(wheat_season, "Wheat: month by month")
```

::: callout-tip
## Checkpoint: what the chart should show

-   AETI is low after sowing (November, December), rises as the crop
    grows and peaks in March and April, then falls as the crop ripens.
-   ETc follows the shape of the Kc curve from Step 3.
-   Rain falls mostly from November to March, so in the first months
    rain can cover most of the consumption.
:::

```{r}
#| label: wheat-monthly-check

m <- monthly_table(wheat_season)
aeti_m <- m[m$variable == "AETI (consumed)", ]
t_m    <- m[m$variable == "T (transpiration)", ]
check(all(t_m$mm <= aeti_m$mm + 1e-6), "monthly T is below AETI in every month")
peak <- format(aeti_m$month[which.max(aeti_m$mm)], "%B")
check(peak %in% c("March", "April"), paste("AETI peaks in", peak))
```

## Step 8: Indicators and maps

```{r}
#| label: wheat-out-aeti-etc
#| fig-cap: "Seasonal AETI and crop water demand (ETc) for wheat"

wheat_etc <- wheat_season$etc_by_class[[as.character(wheat_class_code)]]$etc_seasonal

two_maps(wheat_season$seasonal_aeti$raster, "Seasonal AETI", "mm",
         wheat_etc, "Crop water demand (ETc)", "mm",
         palette1 = "YlGnBu", palette2 = "Oranges")

raster_stats(`AETI (mm)` = wheat_season$seasonal_aeti$raster,
             `T (mm)`    = wheat_season$seasonal_t$raster,
             `ETc (mm)`  = wheat_etc)
```

```{r}
#| label: wheat-out-adequacy-bf
#| fig-cap: "Adequacy and beneficial fraction for wheat"

two_maps(wheat_season$adequacy_etc, "Adequacy", "AETI / ETc",
         wheat_season$beneficial_fraction, "Beneficial fraction", "T / AETI",
         palette1 = "RdYlBu", palette2 = "YlGn", rev1 = FALSE, rev2 = TRUE)

raster_hist(wheat_season$adequacy_etc, "Adequacy (AETI / ETc)", ref = 1)

bf_max <- terra::global(wheat_season$beneficial_fraction, "max", na.rm = TRUE)[1, 1]
check(bf_max <= 1 + 1e-6, sprintf("beneficial fraction never exceeds 1 (max = %.2f)", bf_max))
```

### Irrigation performance indicators

We now assess the wheat fields with the WaPOR-based irrigation
performance framework of Chukalla et al. (2022): **adequacy** (was the
crop water demand met?), **uniformity** (was water use even within a
field?) and **equity** (was water use even between fields?). The
equations and class limits are in @sec-indicators.

::: callout-note
## Concept: fields or blocks

Uniformity is calculated inside a field and equity between fields, so both
need field boundaries. Our crop map is a raster without field boundaries,
so we divide the area into regular **1 km blocks** (50 x 50 pixels of
20 m) and treat each block as a management unit. Blocks with fewer than
100 wheat pixels (4 ha) are left out. With a field or parcel map, use the
fields instead; the method is the same.
:::

**Adequacy classes.** Each wheat pixel is classed as poor, acceptable or
good; pixels above 1 consumed more than the crop demand.

```{r}
#| label: wheat-performance-adequacy
#| fig-cap: "Adequacy classes for wheat (Chukalla et al., 2022)"

adequacy_classes <- terra::classify(
  wheat_season$adequacy_etc,
  rcl = matrix(c(-Inf, 0.68, 1,
                 0.68, 0.80, 2,
                 0.80, 1.00, 3,
                 1.00,  Inf, 4), ncol = 3, byrow = TRUE),
  right = TRUE                                   # classes include their upper limit
)
class_names <- c("Poor (A <= 0.68)", "Acceptable (0.68 to 0.8)", "Good (0.8 to 1)", "Above demand (A > 1)")
levels(adequacy_classes) <- data.frame(value = 1:4, class = class_names)

terra::plot(adequacy_classes, col = c("#d7191c", "#fdae61", "#1a9641", "#2c7bb6"), axes = FALSE,
            main = "Wheat adequacy classes", maxcell = terra::ncell(adequacy_classes))

adequacy_share <- terra::freq(adequacy_classes)
adequacy_share$area_ha  <- round(adequacy_share$count * prod(terra::res(adequacy_classes)) / 10000)
adequacy_share$share_pc <- round(100 * adequacy_share$count / sum(adequacy_share$count), 1)
adequacy_share[, c("value", "area_ha", "share_pc")]
```

::: callout-tip
## Checkpoint

The shares add up to 100%. Low adequacy over large areas is expected
where cereals get little or no irrigation in spring: the demand (ETc) is
calculated for a crop that never lacks water, so rain-fed or deficit
irrigated fields fall in the poor class. Check the pattern with the
irrigation schedule of the scheme before calling it water stress.
:::

**Uniformity and equity.** For each 1 km block we calculate the mean and
the standard deviation of seasonal AETI over its wheat pixels.

$$
Uniformity = 1 - \frac{sd(AETI)}{mean(AETI)} \qquad
Equity = CV = \frac{sd(\text{block means})}{mean(\text{block means})}
$$

```{r}
#| label: wheat-performance-uniformity
#| fig-cap: "Uniformity of water use within 1 km blocks of wheat"

aeti_w     <- wheat_season$seasonal_aeti$raster
block_fact <- 50                                          # 50 x 50 pixels of 20 m = 1 km
block_mean <- terra::aggregate(aeti_w, fact = block_fact, fun = "mean", na.rm = TRUE)
block_sd   <- terra::aggregate(aeti_w, fact = block_fact, fun = "sd", na.rm = TRUE)
block_n    <- terra::aggregate(!is.na(aeti_w), fact = block_fact, fun = "sum", na.rm = TRUE)
use_block  <- block_n >= 100                              # at least 4 ha of wheat

uniformity <- terra::ifel(use_block, 100 * (1 - block_sd / block_mean), NA)
plot_map(uniformity, "Uniformity within 1 km blocks", "%", palette = "RdYlGn", rev = FALSE)

block_means <- terra::values(terra::ifel(use_block, block_mean, NA), na.rm = TRUE)[, 1]
equity_cv   <- 100 * sd(block_means) / mean(block_means)
equity_class <- cut(equity_cv, c(0, 10, 25, Inf), labels = c("good", "fair", "poor"))

data.frame(
  blocks_used         = length(block_means),
  mean_uniformity_pc  = round(terra::global(uniformity, "mean", na.rm = TRUE)[1, 1], 1),
  min_uniformity_pc   = round(terra::global(uniformity, "min", na.rm = TRUE)[1, 1], 1),
  equity_cv_pc        = round(equity_cv, 1),
  equity_class        = equity_class
)
```

::: callout-tip
## Checkpoint

Uniformity is between 0 and 100%. Compare it with the standard for the
irrigation method used in the blocks (drip 85%, sprinkler 75%, furrow
65%); blocks below the standard are where water use inside the fields is
uneven. An equity CV below 10% means water use is shared evenly between
blocks.
:::

**Climate normalization.** Part of the difference in water use between
blocks comes from the weather, not from management: a block with a higher
reference ET (RET) consumes more water under the same management.
Chukalla et al. (2022) remove this effect with a normalization factor:

$$
f_{norm} = \frac{\overline{RET}}{RET_i}
$$

where $\overline{RET}$ is the mean seasonal RET over the area and $RET_i$
the seasonal RET of pixel (or field) $i$. Multiplying AETI or water
productivity by $f_{norm}$ makes them comparable across the area.

```{r}
#| label: wheat-performance-normalization

ret_w  <- wheat_season$seasonal_ret$raster
f_norm <- terra::mask(terra::global(ret_w, "mean", na.rm = TRUE)[1, 1] / ret_w, aeti_w)
raster_stats(`Climate normalization factor` = f_norm)
```

::: callout-tip
## Checkpoint

A factor close to 1 everywhere means the weather is similar over the whole
area, so differences in AETI come mostly from management. Factors that
differ by more than about 10% mean the climate varies across the area,
and AETI or CWP should be normalized before comparing fields.
:::

Next, the water source of the consumed water:

```{r}
#| label: wheat-out-green-blue
#| fig-cap: "Green (rain) and blue (irrigation) water for wheat"

wheat_green <- wheat_season$green_water
wheat_blue  <- wheat_season$blue_water

two_maps(wheat_green, "Green water", "mm", wheat_blue, "Blue water", "mm",
         palette1 = "Greens", palette2 = "Blues")

plot_green_blue_monthly(wheat_season, "Wheat: where the water came from, month by month")

gap <- terra::global(abs(wheat_green + wheat_blue - wheat_season$seasonal_aeti$raster), "max", na.rm = TRUE)[1, 1]
check(gap < 0.01, sprintf("green + blue = AETI on every pixel (largest difference %.4f mm)", gap))

wheat_shortcut <- seasonal_shortcut(wheat_season)
data.frame(
  method   = c("Month by month (correct)", "Seasonal totals (wrong)"),
  green_mm = c(mean_of(wheat_green), mean_of(wheat_shortcut$green)),
  blue_mm  = c(mean_of(wheat_blue),  mean_of(wheat_shortcut$blue))
)
```

::: callout-note
## Concept: why the two methods differ so much for wheat

Winter wheat grows during the rainy season, but the rain is not spread
evenly: December and January are wet while the crop still uses little
water, and in April the crop uses a lot while rain decreases. Split
month by month, April's shortfall shows up as blue water (irrigation or
water stored in the soil). Split on the season totals, the winter
surplus would hide it, and the fields would look almost fully
rain-fed.
:::

```{r}
#| label: wheat-out-productivity
#| fig-cap: "Seasonal biomass and estimated grain yield for wheat"

two_maps(wheat_season$seasonal_biomass_t, "Biomass", "t DM/ha",
         wheat_season$yield_raster, "Grain yield", "t/ha",
         palette1 = "YlGn", palette2 = "YlOrBr")

raster_stats(`Biomass (t/ha)` = wheat_season$seasonal_biomass_t,
             `Yield (t/ha)`   = wheat_season$yield_raster)

data.frame(
  indicator = c("CWP, grain (kg/m³)", "BWP, dry biomass (kg/m³)"),
  value     = round(c(wheat_season$cwp, wheat_season$bwp), 2)
)

# Confirm the yield map follows the yield equation with the factors from Step 3
tbp       <- wheat_season$seasonal_biomass_t                       # TBP = NPP x 22.222 / 1000
y_formula <- wheat_params$HI * wheat_params$AOT * wheat_params$fc * tbp / (1 - wheat_params$MC)
y_gap     <- terra::global(abs(y_formula - wheat_season$yield_raster), "max", na.rm = TRUE)[1, 1]
check(y_gap < 1e-6, sprintf("yield = HI x AOT x fc x TBP / (1 - MC) on every pixel (largest difference %.2g t/ha)", y_gap))
```

::: callout-tip
## Checkpoint

Wheat grain yields in northern Tunisia are typically 2 to 5 t/ha, and
crop water productivity for wheat is usually between 0.6 and 1.7 kg/m³.
For wheat, CWP is lower than BWP because grain is only part of the
biomass.
:::

## Step 9: Save the results

```{r}
#| label: wheat-export

wapor_export_analysis_outputs(
  results    = list(Wheat_Winter2023_2024 = wheat_season),  # the name becomes the sub-folder
  folder     = dirname(wheat_output_dir),
  indicators = wheat_indicators,
  include_dekadal = FALSE       # the dekadal inputs are already saved in wapor_data/
)

wheat_files <- list.files(wheat_output_dir, recursive = TRUE)
check(length(wheat_files) > 0, paste(length(wheat_files), "files written to", wheat_output_dir))
table(folder = dirname(wheat_files))
```

------------------------------------------------------------------------

# Results side by side {#sec-summary}

```{r}
#| label: summary-table

season_summary <- function(season, etc) {
  c(
    `Crop area (ha)`            = round(sum(season$mask_class_stats$area_ha, na.rm = TRUE)),
    `AETI (mm)`                 = mean_of(season$seasonal_aeti$raster),
    `Transpiration T (mm)`      = mean_of(season$seasonal_t$raster),
    `Crop demand ETc (mm)`      = mean_of(etc),
    `Rain (mm)`                 = mean_of(season$seasonal_pcp),
    `Effective rain (mm)`       = mean_of(season$seasonal_peff),
    `Green water (mm)`          = mean_of(season$green_water),
    `Blue water (mm)`           = mean_of(season$blue_water),
    `Adequacy (AETI / ETc)`     = mean_of(season$adequacy_etc),
    `Beneficial fraction`       = mean_of(season$beneficial_fraction),
    `Biomass (t DM/ha)`         = mean_of(season$seasonal_biomass_t),
    `Yield (t/ha)`              = mean_of(season$yield_raster),
    `CWP (kg/m³)`               = num_or_na(season$cwp),
    `BWP (kg/m³)`               = num_or_na(season$bwp)
  )
}

data.frame(
  indicator = names(season_summary(citrus_season, citrus_etc)),
  Citrus    = unname(season_summary(citrus_season, citrus_etc)),
  Wheat     = unname(season_summary(wheat_season, wheat_etc))
)
```

Citrus shows `NA` for biomass, yield, CWP and BWP on purpose: tree crop
yield comes from farm surveys, not from the NPP equation.

```{r}
#| label: crop-comparison
#| fig-cap: "Seasonal water balance per crop (mean over the crop mask)"

balance <- data.frame(
  crop   = rep(c("Citrus", "Wheat"), each = 2),
  source = rep(c("Green (rain)", "Blue (irrigation)"), times = 2),
  mm     = c(mean_of(citrus_green), mean_of(citrus_blue), mean_of(wheat_green), mean_of(wheat_blue))
)

ggplot(balance, aes(crop, mm, fill = source)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("Green (rain)" = "#43a047", "Blue (irrigation)" = "#1e88e5")) +
  labs(x = NULL, y = "Seasonal AETI (mm)", fill = NULL, title = "Where the consumed water came from") +
  theme_minimal()
```

::: callout-warning
## Comparing crops

Citrus is an evergreen tree crop measured over a full year; wheat is
an annual crop measured over seven winter months. Their water
productivity values are **not a ranking** of which crop is more
efficient: season length, climate and the product (fresh fruit versus
dry grain) all differ. Compare each crop with its own benchmarks and
between farms of the same crop.
:::

------------------------------------------------------------------------

# Glossary {.unnumbered}

| Term | Meaning |
|------------------------------------|------------------------------------|
| **Adequacy** | AETI divided by ETc. Below 1: the crop consumed less than its demand. |
| **AETI** | Actual EvapoTranspiration and Interception: all water leaving the land surface as vapour. |
| **AOI** | Area of interest: the outline of the study area. |
| **Beneficial fraction (BF)** | T divided by AETI: the share of consumption used by the crop. |
| **Biomass (DMP)** | Dry matter produced by the crop, from NPP. |
| **Blue water** | Consumption supplied by irrigation (or groundwater). |
| **BWP** | Biomass water productivity: dry biomass per m³ of water consumed. |
| **Crop mask** | Raster that marks which pixels belong to the crop being studied. |
| **CWP** | Crop water productivity: harvested yield per m³ of water consumed. |
| **Dekad** | A period of about 10 days: days 1 to 10, 11 to 20, and 21 to the end of each month. |
| **Effective rain (Peff)** | Part of the rain that stays in the root zone and can be used by the crop. |
| **ETc** | Crop evapotranspiration: water demand of a healthy, well-watered crop, $K_c \times RET$. |
| **Green water** | Consumption supplied by rain. |
| **Harvest index (HI)** | Harvested share of the above-ground biomass. |
| **Uniformity** | Evenness of water use within a field or block: 1 minus the CV of seasonal AETI of its pixels. |
| **Equity** | Evenness of water use between fields or blocks: CV of their mean seasonal AETI. |
| **Climate normalization factor** | Mean seasonal RET of the area divided by the RET of a field; removes the effect of weather when comparing fields. |
| **Kc** | Crop coefficient: ratio between the crop's demand and the reference ET, changing with growth stage. |
| **L1, L2, L3** | WaPOR resolution levels: about 300 m, 100 m and 20 m. |
| **NPP** | Net Primary Production: carbon fixed by the vegetation. |
| **P95** | The 95th percentile: the value exceeded by only the best 5% of pixels. |
| **RET** | Reference evapotranspiration: water demand of a reference grass surface, set by the weather. |
| **T** | Transpiration: water that passes through the plants. |

# References {.unnumbered}

-   Allen, R.G., Pereira, L.S., Raes, D. and Smith, M. (1998). *Crop
    evapotranspiration: Guidelines for computing crop water
    requirements.* FAO Irrigation and Drainage Paper 56. Rome: FAO.
    [Chapter 6, Table
    12](https://www.fao.org/4/x0490e/x0490e0b.htm#chapter%206%20%20%20etc%20%20%20single%20crop%20coefficient%20(kc)).
-   Chukalla, A.D., Mul, M.L., van der Zaag, P., van Halsema, G.,
    Mubaya, E., Muchanga, E., den Besten, N. and Karimi, P. (2022). A
    framework for irrigation performance assessment using WaPOR data: the
    case of a sugarcane estate in Mozambique. *Hydrology and Earth System
    Sciences*, 26, 2759 to 2778.
    <https://doi.org/10.5194/hess-26-2759-2022> (PDF:
    <https://hess.copernicus.org/articles/26/2759/2022/hess-26-2759-2022.pdf>).
    Source of the equations and class limits for adequacy, uniformity,
    equity, land and water productivity, and the climate normalization
    factor.
-   FAO WaPOR data portal: <https://data.apps.fao.org/wapor>
-   USDA Soil Conservation Service (1970). Irrigation Water
    Requirements. Technical Release No. 21 (effective rainfall method).
